Stimulatory and inhibitory G-protein signaling relays drive cAMP accumulation for timely metamorphosis in the chordate Ciona

Version 1

Abstract

Abstract

Larvae of the ascidian Ciona initiate metamorphosis tens of minutes after adhesion to a substratum via its adhesive organ. The gap between adhesion and metamorphosis initiation is suggested to ensure the rigidity of adhesion, allowing Ciona to maintain settlement after losing locomotive activity through metamorphosis. The mechanism producing the gap is unknown. Here, by combining gene functional analyses, pharmacological analyses, and live imaging, we propose that the gap represents the time required for sufficient cAMP accumulation to trigger metamorphosis. Not only the Gs pathway but also the Gi and Gq pathways are involved in the initiation of metamorphosis in the downstream signaling cascade of the neurotransmitter GABA, the known initiator of Ciona metamorphosis. The mutual crosstalk of stimulatory and inhibitory G-proteins functions as the accelerator and brake for cAMP production, ensuring the faithful initiation of metamorphosis at an appropriate time and in the right situation.

Introduction

Metamorphosis is a widespread feature of animal development that allows them to have different functions between larval and adult stages 1. As larvae become adults, the shape and various characteristics of their physiology, gene expression, behavior, and lifestyle change. With these changes, animals can focus on a limited number of biological activities at each stage to increase feeding, growth, dispersal, and reproduction efficiencies. This biphasic mode of life is a widely conserved and ancient trait of animals, as evidenced by its presence in groups that maintain primitive states 2, indicating that this long-conserved feature has contributed to animals’ flourishment. Therefore, to understand animal evolution, it is essential to elucidate the mechanisms underlying metamorphosis.

A key characteristic of metamorphosis is that both internal and external conditions determine its initiation. Sufficiently matured (or metamorphically competent3) larvae start metamorphosis only when they meet an appropriate external condition to be adults. Various organic and inorganic external stimuli suited to the lifestyle of adulthood trigger metamorphosis. Many marine invertebrates exhibit a benthic lifestyle at the adult stage4. Their planktonic larvae have an adhesive organ that secretes adhesives and triggers metamorphosis when they adhere to a substratum. Upon starting metamorphosis, their larvae lose locomotive organs and transition into benthic adult forms. Determining the timing of metamorphosis is important for benthic animals, particularly sessile ones, to ensure their future survival and reproduction, as they become unable to move after metamorphosis. Therefore, sessile animals are suspected to have elaborate mechanisms to start metamorphosis only when a firm adhesion is achieved at an appropriate place. Although several reports describe this kind of phenomenon5–7, the mechanism by which the external condition is turned into an internal mechanism to trigger metamorphosis at the right timing remains elusive.

Ascidians are marine invertebrate chordates that are the closest living relatives to vertebrates8,9. The chordate features of ascidians are best represented by their tadpole larval shape. Like vertebrate tadpoles, ascidian larvae swim by beating their tail through neuromuscular activities10,11. However, ascidians lose the tadpole shape during metamorphosis, and they exhibit a sessile lifestyle at the adult stage12,13. Ascidian metamorphosis comprises several key events, such as tail regression, body axis rotation, and adult organ growth14 (Figure 1A).

Gq and Gs pathways are required for Ciona metamorphosis.

(A) Schematic illustration of Ciona metamorphosis.

(B) A larva knocked down with Gαq using the antisense morpholino oligonucleotide (MO). Metamorphosis did not initiate at 2 days post-fertilization (2 dpf).

(C) A Gαs knockdown larva.

(D) A control animal injected with the standard (STD) MO. Metamorphosis initiated, as indicated by the completion of tail regression.

(E) Effect of Gαq knockdown on the percentage of metamorphosis initiation (indicated by the initiation of tail regression), shown as box-and-whisker plots. Dots indicate experiment replicates. ***, p<0.001 (Fisher’s exact test). n, number of examined larvae in total.

(F) Effect of Gαs knockdown.

(G-I) Effects of PLCβ1/2/3, PLCβ4, and IP3R knockdowns.

(J-M) Results of the experiments where adhesion was prevented by tail amputation and laying larvae on agar-coated plates.

(J) A control larva overexpressing the Kaede reporter in the entire nervous system using the PC2 cis element.

(K) A larva overexpressing constitutive active (ca)PLCβ1/2/3.

(L) Effect of caGαq overexpression.

(M) Effect of caPLCβ1/2/3 overexpression.

(N-O) Live imaging of Ca2+ transient, monitored by injecting GCaMP8 mRNA.

(N) Snapshot of the control larva injected with GCaMP8 mRNA. An increase in the fluorescence intensity in the adhesive papillae (arrow) was observed. Rod, the position of the glass rod used to stimulate papillae. The percentage and number exhibit the rate of animals showing Ca2+ transient in the papillae.

(O) A larva injected with Gαq MO plus GCaMP8 mRNA. An increase in the fluorescence intensity in the papillae (arrow) was not observed even though the brightness of the whole body was raised. See also Figure S1.

Adhesion to a substratum triggers ascidian metamorphosis14. Ascidian larvae have adhesive organs called adhesive papillae, usually triangular protrusions at the anterior end. Adhesive papillae secrete a lectin-type mucus substance that is suspected to aid in binding to the substratum15. Moreover, adhesive papillae serve as mechanical sensors16. The papilla is a neuronal organ that includes epidermal neurons positive for vesicular glutamate transporter (vGLUT) and GABA, which innervate axons toward the sensory vesicle (larval brain)17–20. The papilla neurons are responsible for initiating metamorphosis16,21–23. Moreover, our recent study suggests the involvement of a mechanoreceptor channel (the TRP channel) in the responsiveness to mechanical stimuli22. However, a simple physical adhesion is not sufficient to trigger metamorphosis. In our study using the model ascidian Ciona intestinalis Type A (this species has been suggested to be renamed C. robusta 24–26, and hereafter we call it Ciona), continuous adhesion for about 30 min is necessary before starting metamorphosis27. When larvae detach before reaching the critical period, they must adhere for 30 min again for metamorphosis. Therefore, ascidian larvae are suspected of sensing the duration of adhesion and initiating metamorphosis only when adhesion is firm enough to be maintained for the required period. Ciona larvae continue tail beating during adhesion to push their body to the substratum. We showed that the strength of force generated by this swimming activity influences the timing of metamorphosis initiation22. In this phenomenon, abolishment of swimming activity elongates the time from settlement until metamorphosis is initiated. The mechanisms for measuring the duration of adhesion and the strength of the force generated by adhesion are unknown.

To elucidate the mechanisms triggering ascidian metamorphosis, the signaling pathways must be characterized. Many studies have discovered signaling molecules as possible inducers of ascidian metamorphosis28–35. These molecules include neurotransmitters, suggesting that transmission of excitation in the nervous system, starting from the adhesive papillae, is crucial for metamorphosis. Recently, our group reported that the inhibitory neurotransmitter GABA plays a pivotal role in initiating Ciona metamorphosis36. Knockout and knockdown of the genes encoding GABA synthase (GAD), vesicular inhibitory amino acid transporter (VIAAT or VGAT), and metabotropic GABA receptor (GABABR) resulted in the perturbation of metamorphosis. Moreover, GABA administration induces metamorphosis without adhesion. Our previous studies demonstrated that GABA induces the secretion of gonadotropin-releasing hormone (GnRH)36,37; however, it remains unknown how the inhibitory neurotransmitter activates neuronal functions for initiating metamorphosis.

In this study, we addressed the characterization of the downstream cascade stimulated by GABA. Because GABABR is a G-protein-coupled receptor (GPCR)38, we searched for trimeric G-proteins necessary for metamorphosis. These G-proteins are heterotrimers of an α β and γ subunit39. Upon ligand binding, GPCR exchanges GDP of the α subunit to GTP, then the GTP-bound α subunit and the βγ complex are released from the GPCR. Although both the GTP-bound α subunit and the βγ complex have activities, the α subunit mainly determines the reaction specific to the G-protein type. We found that three G-proteins are activated in the downstream GABA cascade, resulting in cyclic adenosine monophosphate (cAMP) elevation. Because the signaling includes stimulating and inhibiting cAMP synthesis, Ciona initiates metamorphosis only when a sufficient quantity of cAMP is accumulated due to sustained adhesion. Our results revealed the ingenious mechanism that permits Ciona to start metamorphosis only when achieving an appropriate adhesion firm enough to relinquish swimming ability and commence sessile adult life.

Results

Characterization of G-proteins necessary for metamorphosis initiation

We knocked down the genes encoding G alpha (Gα) proteins40 using antisense morpholino oligonucleotides (MOs). We found that two genes corresponding to Gαq and Gαs are necessary for metamorphosis. Neither morphant showed any signature of metamorphosis even though both were allowed to adhere to the base of culture dishes, which was sufficient for control larvae to initiate metamorphosis (Figure 1A-F).

Gαq activates phospholipase C beta (PLCβ) to produce inositol triphosphate (IP3)41. IP3 is received by its receptor on the endoplasmic reticulum (ER) and releases calcium ion (Ca2+). The Ciona genome encodes two PLCβ (PLCβ1/2/3, PLCβ4) and one IP3 receptor (IP3R) (Figure S1 and Table S1). We knocked down PLCβ1/2/3, PLCβ4, and IP3R genes. The knockdown larvae of these three genes failed to start metamorphosis (Figure 1G-I), suggesting that Gαq initiates metamorphosis by the conventional Ca2+ pathway mediated by PLCβ and IP3/IP3R. To confirm this, we overexpressed a constitutively active form of Gαq (caGαq)42 and of caPLCβ1/2/343 in the entire nervous system with the cis element of the gene encoding prohormone convertase 2 (PC2)44–46. When the microinjected animals reached the larval stage, they were cultured on agar-coated dishes after tail amputation to prevent adhesion. In this condition, the control larvae rarely started metamorphosis because of the absence of adhesion (Figure 1J). In contrast, caGαq- or caPLCβ1/2/3-overexpressed larvae initiated metamorphosis without adhesion (Figure 1K-M).

Adhesive papillae exhibit a Ca2+ increase soon after sensing an adhesive stimulus16,22. We examined whether this Ca2+ transient is dependent on the Gq pathway. Compared to controls, significantly fewer larvae injected with Gαq MO plus GCaMP8 mRNA exhibited GCaMP8 fluorescence elevation in the adhesive papilla upon stimulation (Figure 1N, O). This result suggests that the Gq pathway is activated upon adhesion to cause a Ca2+ transient in the adhesive papilla.

Gs pathway initiates metamorphosis by activating cAMP synthesis

The involvement of Gs in metamorphosis was confirmed by the overexpression of a constitutively active form of Gαs (caGαs)47 in the nervous system. This overexpression resulted in the initiation of metamorphosis without adhesion (Figure 2A-C). The Gs pathway activates adenylate cyclase (AC) to produce cAMP48. We previously reported that cAMP can induce metamorphosis37. Because externally added cAMP is not a strong inducer of metamorphosis, we attempted to confirm this hypothesis through another experiment. Theophylline increases cAMP by inhibiting the cAMP-degrading enzyme phosphodiesterase (PDE)49. We treated wild-type larvae with theophylline after tail amputation to prevent larvae from adhesion. Most theophylline-treated larvae completed tail regression without adhesion (Figure 2D-F). Theophylline has several target proteins in addition to PDE50. To further confirm that cAMP is responsible for the initiation of metamorphosis, we overexpressed photo-activating AC (bPAC)51 in the nervous system. The bPAC-overexpressed larvae regressed their tails without adhesion, suggesting that a cAMP increase triggers metamorphosis (Figure 2G).

cAMP initiates Ciona metamorphosis.

(A) An adhesion-prevented control (uninjected) larva at 2 dpf. Metamorphosis did not initiate.

(B) An animal with adhesion prevention and overexpression of caGαs in the entire nervous system. Metamorphosis initiated.

(C) Effect of caGαs overexpression.

(D) A control (uninjected) larva.

(E) A theophylline-treated animal. Metamorphosis initiated.

(F) Effect of theophylline treatment.

(G) Effect of bPAC overexpression in the nervous system.

(H) Effect of bPAC overexpression over time. In this experiment, adhesion was permitted. Control groups (overexpressed with mCherry) exhibited a slower initiation of metamorphosis than bPAC-overexpressed groups.

Using bPAC, we addressed whether enhanced cAMP production facilitates the initiation of metamorphosis. At 24 hours post-fertilization (hpf), only a few animals in both the bPAC-overexpressed and control groups initiated metamorphosis because 24 hpf is somewhat too early for Ciona larvae to be metamorphically competent27 (Figure 2H). Even in this condition, the bPAC-overexpressed group exhibited a statistically higher rate of metamorphosis-initiated larvae. The proportion of animals initiating metamorphosis increased over successive time points, with the bPAC-overexpressed group consistently showing a statistically higher rate of metamorphosis compared to controls. These results support the idea that cAMP plays a role as a timer for Ciona metamorphosis; accumulation of this molecule to exceed a threshold could trigger metamorphosis. Because many of the bPAC-overexpressed larvae did not initiate metamorphosis at 24 hpf, similar to control larvae, cAMP accumulation does not alter the timing of acquiring metamorphic competence.

Gq-Gs pathways work in the adhesive papillae for metamorphosis

The above results showed that activation of the Gq and Gs pathways are the key events in initiating metamorphosis. Gαq is necessary to induce Ca2+ transients in the adhesive papillae, suggesting that the Gq pathway functions in this region. If both Gq and Gs function in the papillae, they should be expressed in the adhesive papilla. The transcriptome analysis of the larval papilla region showed that the genes encoding Gq and Gs pathway proteins are expressed in this region (Figure S2 and Table S1). Therefore, Gq and Gs could function in the papilla to initiate metamorphosis.

As shown above, theophylline induced metamorphosis without settlement. However, when papillae were removed from larvae, theophylline failed to induce metamorphosis (Figure 3A-C). This suggests that cAMP elevation in the adhesive papillae is essential for starting metamorphosis. The overexpression of caGαq, caPLCβ1/2/3, caGαs, and bPAC by the PC2 cis element resulted in the initiation of metamorphosis without adhesion. Like theophylline, amputation of the papillae inhibited them from starting metamorphosis (Figure 3D), confirming that activation of the Gq and Gs pathways in the adhesive papillae triggers metamorphosis.

Gq and Gs pathways are activated in the adhesive papillae to initiate metamorphosis.

(A) A theophylline-treated animal. Metamorphosis initiated.

(B) A papilla-amputated larva treated with theophylline. Metamorphosis did not initiate.

(C) Effect of papilla amputation on theophylline treatment.

(D) Effects of papilla amputation on caGαq, caPLCβ1/2/3, caGαs, and bPAC overexpression.

(E) Measurement of cAMP concentration, as indicated by the fluorescence of Pink Flamindo. Left, the quantification of Pink Flamindo fluorescent intensity is shown for a larva that initiated metamorphosis at the time indicated by an arrowhead. The time of stimulation of the adhesive papillae with a glass rod is denoted by a bar. Right, the position of the region of interest (ROI). See also Figure S3.

If the Gs pathway is activated in the adhesive papillae, a cAMP increase should be observed in this region upon adhesion. We examined this possibility using a fluorescent cAMP indicator called Pink Flamindo (PF)52. After stimulation, the fluorescence of PF in the papillae temporarily decreased on average by 0.83-fold (n=5) from the initial intensity, followed by a gradual increase to 1.17-fold over successive time points (Figure 3E). Such an increase in fluorescence intensity was not observed in the adhesive papillae of the larvae that had failed to initiate metamorphosis following stimulation (n=4, Figure S3). Therefore, increased cAMP in the papillae serves to indicate that sufficient stimuli of adhesion have been received to induce metamorphosis. This strengthens our hypothesis that cAMP accumulation in the adhesive papillae determines the initiation of metamorphosis.

GABA in the adhesive papillae is responsible for metamorphosis

Our previous study demonstrated that GABA is the chief neurotransmitter that induces metamorphosis36. To gain insight into the relationships between the GABA, Gq, and Gs pathways, we addressed whether GABA functions in the adhesive papillae for initiating metamorphosis, similar to Gq and Gs.

The previous studies and our transcriptome data suggest that GAD is expressed in the adhesive papillae53 (Table S1). Moreover, GABA-immunopositive signals have been detected in the papillae19,53. Therefore, papillae can provide GABA to stimulate themselves. Next, we examined whether the adhesive papillae can receive GABA to initiate metamorphosis. We showed that the metabotropic GABA receptor is responsible for the initiation of metamorphosis36. Using whole-mount in situ hybridization, the previous study did not detect the expression of two GABAB receptor (GABABR) genes in the papillae53; however, our transcriptome data detected the low-level expression of three genes encoding GABABR (Table S1) in the papillae, suggesting that adhesive papillae could receive GABA. GABA can induce metamorphosis without adhesion (Figure S4A)36. However, amputation of adhesive papillae eliminated this activity (Figure S4B-C), suggesting that GABA reception by the papillae is responsible for starting metamorphosis.

The larval brain is the major domain expressing GABABR53. Therefore, it remains possible that GABA signaling in the brain stimulates the papillae in a retrograde manner to initiate metamorphosis. The connectome analyses of the larval nervous system did not suggest a nerve that inputs into the papillae from the brain54,55; however, retrograde stimulation would be possible through the volume transmission of a signaling molecule. To examine this possibility, we isolated larval fragments anterior to the brain (Figure S4D), followed by the administration of GABA and theophylline. Through the application of these chemicals, anterior fragments exhibited increased clarity, elongation, and retraction of papillae, which are the signatures of metamorphosis (Figure S4E-G). Because their responses to GABA were weaker than those observed in the theophylline treatment, we further examined whether the anterior fragments can respond to GABA by monitoring the expression of GABA-responsive genes. By comparing expression levels between control and GABA-administered larvae, we made a list of the genes upregulated by GABA in the papillae (Table S2). Among them, we compared the expression levels of two genes (KY21.Chr5.240 and KY21.Chr8.489) between GABA-administered and control anterior fragments. GABA-treated anterior fragments expressed these genes at levels at least three times higher than controls (Figure S4H). We concluded that GABA is secreted from and stimulates the papillae to start metamorphosis.

Gs/cAMP is downstream of the Gq pathway

Our next question is: What are the upstream or downstream relationships between GABA, Gq/Ca2+, and Gs/cAMP in the adhesive papillae? We first examined the relationships between the Gq and Gs pathways. Theophylline ameliorated metamorphosis-failed phenotypes in Gαq, PLCβ, and IP3R knockdowns (Figure 4A-E). Moreover, Gαq morphants initiated metamorphosis when caGαs was overexpressed in the nervous system (Figure 4F). These results suggest that the Gq pathway is upstream of the Gs/cAMP pathway. Gαs knockdown larvae exhibited Ca2+ transients in the adhesive papillae upon stimulation (Figure 4G). Because this Ca2+ transient is Gq-dependent (Figure 1O), this result confirms that the Gs pathway is downstream of the Gq-dependent Ca2+ increase.

Relationships between Gq and Gs pathways in metamorphosis.

(A) A Gαq knockdown larva.

(B) Theophylline ameliorated the effect of Gαq knockdown.

(C) Effect of theophylline on Gαq knockdown.

(D) Effect of theophylline on PLCβ knockdown. In this experiment, PLCβ1/2/3 and PLCβ4 were knocked down simultaneously.

(E) Effect of theophylline on IP3R knockdown.

(F) Effect of caGαs overexpression on Gαq knockdown.

(G) Gαs is not necessary for Ca2+ transient in the adhesive papillae (arrow) upon adhesion.

(H) A Gαq knockdown and Kaede-overexpressed larva.

(I) A Gαq knockdown and caPLCβ1/2/3-overexpressed larva.

(J) A Gαs knockdown and caPLCβ1/2/3-overexpressed larva.

(K) Effect of caPLCβ1/2/3 overexpression on Gαq knockdown.

(L) Effect of caPLCβ1/2/3 overexpression on Gαs knockdown.

(M) Effect of caGαq overexpression on Gαs knockdown.

(N) Effect of theophylline treatment on Gαs knockdown.

To gain further insight into the epistatic order of the Gq and Gs pathways, we overexpressed constitutive active forms of Gq pathway proteins in Gαs morphants with the PC2 cis element. If Gq is upstream of the Gs pathway, forced activation of the Gq pathway by caGαq or caPLCβ1/2/3 would not ameliorate the Gαs MO effect. Indeed, caPLCβ1/2/3, which rescued Gαq morphants well, failed to rescue Gαs morphants (Figure 4H-L). In contrast, caGαq significantly ameliorated the metamorphosis-failed phenocopies of Gαs morphants (Figure 4M), contradicting the hypothesis that Gq is upstream of the Gs pathway. One possibility is that the Gq pathway stimulates cAMP synthesis through a massive Ca2+ increase and protein kinase C (PKC) activation56,57. Indeed, Ciona larvae can synthesize cAMP without Gs, as suggested by the partial rescue of Gαs morphants by theophylline treatment (Figure 4N). These data suggest that Gαs-independent AC could be a target of the Gq pathway. We concluded that the Gq pathway is upstream of the Gs pathway in the signaling cascade initiating Ciona metamorphosis.

Interaction between GABA and Gq pathways

We next investigated the relationships between the GABA and Gq/Gs pathways. As our previous study showed, GABA pathway knockdown by GAD or GABABR1 MO disrupted the initiation of metamorphosis (Figure 5A)36. Overexpression of caPLCβ1/2/3 and theophylline treatment ameliorated the metamorphosis-failed phenocopies of GAD/GABABR1 morphants (Figure 5B-D). These results suggest that the GABA pathway is upstream of the Gq and Gs pathways. We observed Ca2+ transients in the adhesive papillae of GAD knockdown larvae. GAD morphants rarely exhibited a Ca2+ increase after stimulation (Figure 5E). Because the Ca2+ transient in the papillae is Gq-dependent, this result confirms that the GABA pathway is upstream of, or at least in parallel with, the Gq pathway.

GABA functions in the adhesive papillae to induce metamorphosis.

(A) A GABABR1 knockdown and mCherry-overexpressed larva.

(B) A GABABR1 knockdown and caPLCβ1/2/3-overexpressed larva. Metamorphosis initiated.

(C) Effect of caPLCβ1/2/3 overexpression on GABABR1 knockdown.

(D) Effect of theophylline on GABABR1 and GAD knockdowns.

(E) GABA is necessary for Ca2+ transient in the adhesive papillae (arrow).

(F) Effect of GABA on Gαq knockdown.

(G) Effect of GABA on PLCβ knockdown.

(H) Effect of GABA on Gαs knockdown. See also Figure S4.

However, puzzling results were obtained when we administered Gαq- and Gαs-knockdown larvae with GABA. If GABA is upstream of the Gq and Gs pathways, this chemical does not induce metamorphosis when the Gq or Gs pathway is disrupted. However, GABA significantly ameliorated the metamorphosis-failed phenocopies of Gαq, PLCβ, and Gαs morphants (Figure 5F-H). These results could be explained by assuming enhancement of the Gq pathway by GABA through PLCβ and another GABA-mediated metamorphic pathway bypassing Gq components. These possibilities are examined in the next section.

Contribution of Gi to metamorphosis

The function of GABA as the potential upstream factor of Gq could be easily explained if GABABR is coupled with Gq. Although a few studies suggest this possibility58, Gi is regarded as the major G-protein coupled with GABABR59. A previous study suggested that the Ciona genome encodes one typical Gαi protein40. The gene encoding this Gαi is strongly expressed in the entire larval nervous system including adhesive papillae60. The knockdown of this gene exhibited a significantly reduced rate of metamorphosis (Figure 6A); however, the reduction was not conspicuous, suggesting the presence of another Gαi regulating metamorphosis.

Gi is required for metamorphosis initiation.

(A) Effect of Gαi knockdown.

(B) A control animal injected with STD MO.

(C) A dvGαi_Chr2 knockdown larva.

(D) A dvGαi_Chr2 knockdown and theophylline-treated animal.

(E) Effects of GABA and theophylline on dvGαi_Chr2 knockdown.

(F) Effects of wtGαi and wtGαs overexpression.

(G) Effect of dnGαi overexpression.

(H) Effect of GABABR1 and Gαq simultaneous knockdown.

(I) Effects of caPLCβ1/2/3 and caGαs overexpressions on GABABR1 and Gαq simultaneous knockdown.

(J) GIRK knockdown promoted metamorphosis initiation.

(K) Effect of U0126 on GABA treatment.

(L) Effect of U0126 on theophylline treatment. See also Figure S5.

A previous study showed that Ciona has four Ciona-specific Gα proteins40. We found three gene models (KY21.Chr2.875, KY21.Chr4.943, and KY21.Chr9.455) encoding these divergent Gα proteins from the newest genome database61. Our phylogenetic analyses showed that the divergent Gα proteins have a strong affinity to the Gαi/o family proteins (Figure S5A). Moreover, the amino acid residues at the C-terminal end of KY21.Chr2.875 are more similar to human Gαi than the other Gα proteins (Figure S5B). The C-terminal residues, particularly glycine at position 3, are essential for determining the G-protein partner of GPCR62. We tentatively name this protein dvGαi_Chr2. The transcriptome data showed that the gene encoding dvGαi_Chr2 is strongly expressed in the papillae (Table S1). The dvGαi_Chr2 knockdown larvae failed to initiate metamorphosis (Figure 6B-C), and this phenocopy was rescued by theophylline and GABA (Figure 6D-E), suggesting that dvGαi_Chr2 is necessary for metamorphosis. This strengthens the hypothesis that GABABR regulates metamorphosis through Gi. The rescue of dvGαi_Chr2 morphants by GABA could be attributable to a redundant function of the typical Gαi protein in the papillae.

We further explored the involvement of Gi upon metamorphosis as the upstream factor of the Gq-Gs pathways. Gi is known to activate PLCβ through the Gβγi complex63,64. Released βγ is inactivated by overexpressing normal (usually GDP-bound) Gα subunits because GDP-Gα quenches the βγ complex65. Overexpressing wild-type Gαi and Gαs in the nervous system suppressed metamorphosis (Figure 6F), suggesting that the activated βγ complex is necessary for initiating metamorphosis. To confirm that the wild-type Gαi exerts its effect through the sequestration of the βγ complex, we overexpressed a dominant-negative form of Gαi (dnGαi)66,67 which has reduced GDP/GTP affinity while maintaining βγ binding activity. dnGαi significantly reduced the occurrence of metamorphosis (Figure 6G), suggesting that the negative effect of Gαi on metamorphosis occurs through interaction with the βγ complex.

If Gβγ-mediated PLCβ activation is used in Ciona metamorphosis, PLCβ receives two independent inputs (Gαq and Gβγi) for its activation. These pathways could compensate for each other. We noticed that the MOs for GABA pathway genes and the Gαq MO did not strongly disrupt metamorphosis compared to the Gαs MO (Figure 1E-F; 5C-D). The compensatory role could explain this phenomenon. Indeed, the simultaneous knockdown of GABABR1 and Gαq resulted in a strong impairment of metamorphosis (Figure 6H), and these morphants initiated metamorphosis by overexpression of caPLCβ1/2/3 or caGαs (Figure 6I).

If the role of the GABA/Gi pathway relies specifically on PLCβ activation in the mechanism of metamorphosis, GABA administration would not rescue PLCβ morphants. However, GABA weakly but significantly induced metamorphosis in PLCβ1/2/3 plus PLCβ4 double-knockdown larvae (Figure 5G). Therefore, GABA/Gi is likely to activate another pathway that bypasses PLCβ. Gβγi is known to activate group III AC56. Its Ciona counterpart (AC5/6) is expressed in the adhesive papillae (Figure S2B), which could explain the results of the rescue experiments. In addition, there are two other major targets of Gβγi. One is the G-protein-activated inwardly rectifying potassium (GIRK) channel68. Our RNA-seq data on the papillae indicated the expression of two GIRK channel genes (Table S1). The GIRK channel negatively regulates the excitation of neurons through hyperpolarization. If the metamorphosis of Ciona is induced by the excitation of papilla neurons as suggested previously16,22, the GIRK channel is likely to regulate metamorphosis negatively. Indeed, the knockdown of one GIRK channel gene weakly enhanced the initiation of metamorphosis without settlement (Figure 6J). Although the negative role of the GIRK channel supports the involvement of Gβγi in the pathway of metamorphosis, this does not explain the PLCβ-independent activation of the metamorphic pathway by GABA/Gi.

The third function of Gβγi is to activate MAPK signaling through positive regulation of MEK69. PLCβ does not mediate this pathway. It has been suggested that MAPK signaling is required to induce metamorphosis70,71. To show that MEK activation is necessary for inducing metamorphosis mediated by GABA, we treated adhesion-prevented larvae with GABA plus U0126, a potent MEK1/2 inhibitor repeatedly used in ascidians72,73. U0126 reduced the rate of metamorphosis induction by GABA (Figure 6K).

cAMP is also known to activate the pathway that involves MEK1/274. We found that U0126 antagonized the effect of theophylline on metamorphosis (Figure 6L). Therefore, GABA and cAMP have the same target (MEK1/2) to initiate metamorphosis, suggesting their compensatory function and/or that cAMP synthesis is upregulated by GABA-Gβγi. These bypassing pathways could explain the amelioration of the phenocopies of PLCβ and Gαs morphants by GABA (Figure 5G and H).

The constitutive function of wild-type Gαq

We found that Ciona wild-type Gαq (wtGαq) has constitutive activity. In contrast to wtGαs and wtGαi, overexpression of wtGαq did not arrest metamorphosis (Figure S6A). Rather, its overexpression induced metamorphosis without settlement (Figure S6B). The constitutive wtGαq activity was confirmed by the rescue of Gαs morphants (Figure S6C). As mentioned above, caGαq also has these activities (Figure 1L). caGαq-overexpressed larvae had a somewhat rounder trunk shape, abnormal tail bending, and frequent tail twitching, perhaps due to a constitutive increase in the cytosolic Ca2+ concentration (Figure S6D). Overexpression of wtGαq did not show such abnormalities (Figure S6E), suggesting that wtGαq has milder activity than, or works through a different mechanism from, caGαq. To gain further insight into the constitutive effect of wtGαq, we constructed a dominant-negative form of Gαq66 that has reduced GDP/GTP binding activity while maintaining βγ binding activity. Overexpression of dnGαq weakly inhibited metamorphosis (Figure S6F), suggesting that GDP/GTP exchange is important for the constitutive activity of wtGαq.

Discussion

Among chordates, the tunicate ascidian is the only group that exhibits a sessile lifestyle at the adult stage. Understanding how ascidians acquired the mechanism to metamorphose into sessile adults is important for elucidating the evolution of chordates75,76. Through extensive molecular, physiological, and pharmacological analyses, we identified signaling molecules responsible for the initiation of metamorphosis of the ascidian Ciona. Combining the results with previous knowledge, we described a schematic that best represents the cascades running in the adhesive papillae upon adhesion to initiate metamorphosis (Figure 7). This working hypothesis will be the basis for future research on ascidian/tunicate metamorphosis. The hypothesis explains the characteristics of Ciona metamorphosis, the mechanisms of which have not yet been elucidated. The genes, proteins, and signaling pathways in this schematic are essential targets for elucidating how the ancestor of ascidians acquired the metamorphosis system peculiar to this group during evolution.

Schematic illustration of the signaling cascades for initiating Ciona metamorphosis. AC, adenylate cyclase; ER, endoplasmic reticulum; GPCR, G-protein coupled receptor; PKA, protein kinase A.

Mechanisms measuring duration and strength of adhesions

One mystery of Ciona metamorphosis is that its initiation requires continuous adhesion with the adhesive organ for a certain period. This requirement is suggested for the faithful achievement of metamorphosis only when adhesion is firm enough to allow the transition out of the free-living larval stage. Our previous study suggested that approximately 30 min of adhesion is necessary27. A shorter duration of adhesion does not trigger metamorphosis. Larvae that experience short-term adhesion (such as detaching before 30 min) need to adhere again for 30 min to initiate metamorphosis, suggesting that the experience of temporal adhesion is erased. Our recent study showed that the force given to the papillae by adhesion also affects metamorphosis initiation: weaker force extends the time requirement22. Therefore, Ciona larvae likely possess a mechanism that somehow measures the duration and strength of adhesion and the ability to cancel transient excitations of the adhesive organ. Because fewer than 300 neurons constitute the larval nervous system of Ciona54,55,77, how the larva measures the strength and duration of adhesion with its simple nervous system is an important question to address the mechanisms of metamorphosis.

This study showed that cAMP is an essential second messenger molecule for triggering metamorphosis. Because PDE constitutively degrades cAMP, accumulation of this molecule requires persistent and/or strong activation of the metamorphic pathway that overcomes PDE activity. In other words, the duration and strength of adhesion could be converted into the quantity of cAMP, and when its quantity reaches a threshold, metamorphosis can be initiated. The time between the start of adhesion and metamorphosis initiation may be the period necessary to accumulate sufficient cAMP. This “cAMP timer” mechanism does not demand a complicated neuronal network system and could work in the simple nervous system of Ciona.

G-protein signaling relay for metamorphosis

This study showed that three major G-proteins are involved in the initiation of metamorphosis (Figure 7). This complicated mechanism contrasts the simple determination of metamorphosis initiation by cAMP accumulation. What factor promoted the ancestor of ascidians to acquire this complicated signaling network of metamorphosis? We suspect that incorporating multiple components increases the chance of generating the crosstalk between molecules, providing rigidity and flexibility to the system. Particularly, the involvement of Gi may have been important for achieving metamorphosis at the correct time and condition. GABA and Gi have inhibitory and excitatory functions78,79. As we discussed, Ciona larvae must ignore transient adhesion or stimulation on the papilla, which may occur frequently during swimming by colliding with an object or by strong water flows. At the same time, larvae must maintain sensing ability and start metamorphosis when they encounter an appropriate stimulus. Through the GIRK channel and Gαi, GABA can suppress the excitation and cAMP accumulation in the papilla. The GIRK channel could increase the excitation threshold of papilla neurons, allowing them to ignore weak stimuli. The transient reduction of cAMP (Figure 3E) after papilla stimulation guarantees the requirement of long-time adhesion irrespective of the initial quantity of cAMP. This cAMP reduction could be partly explained by the inhibitory function of Gαi on ACs. PDE1 is known to be activated by Ca2+/calmodulin80. Because PDE1 is abundantly expressed in the papillae, the increase in Ca2+ upon adhesion could also trigger sudden cAMP reduction through PDE1.

Together with its inhibitory function, GABA can activate the Gq pathway, which promotes metamorphosis through Gβγi. Group III AC and MEK could also be the targets of Gβγi. When adhesion is long enough, their activations somehow overcome the inhibitory functions of GABA. The GABA pathway is an ideal player that fulfills multiple requirements in the regulation of metamorphosis through its excitatory and inhibitory functions. We suspect that papillae themselves are the source of GABA for metamorphosis initiation. Because VIAAT/VGAT is not expressed in the papillae, the secretion of GABA from the papilla neurons requires an atypical mechanism. GABA is detected in the cell body of the neurons19. The release of cytosolic GABA may be triggered by the reception of mechanical stimuli using the TRP channel22. A VIAAT-independent cytosolic GABA release is observed in pancreatic beta cells81. Curiously, papilla neurons express transcription factor Islet82.

GPCRs for initiating metamorphosis and atypical Gαq activity

Future studies need to specify the mechanisms underlying Gαi, Gαq, and Gαs activation as well as the interaction between them more precisely. The activation of trimeric G-proteins relies on GPCRs, suggesting the presence of the receptors coupled with Gi, Gq, and Gs for metamorphosis initiation (Figure 7). Among them, Gi is likely to be coupled with GABABR; however, the expression of GABABR in the papillae is weak, and their interaction needs to be clarified in the future. We found that wild-type Gαq exhibits constitutive activity that does not require activation through adhesion. The constitutive Gαq activity requires GTP binding. If a GPCR owns this Gαq activation, its ligand should be supplied constitutively. Wild-type Ciona does not initiate metamorphosis without adhesion even though Gαq is expressed in the adhesive papillae, suggesting that the quantity of Gαq is strictly regulated to prevent autonomous metamorphosis in normal conditions.

We did not identify the GPCR coupled with Gs in the metamorphic mechanism. We suspect that GnRH receptors (GnRHRs) are strong candidates for this role. Our previous studies showed that GnRHs can induce metamorphosis36,37 as a downstream factor of GABA. Among the four genes encoding GnRHR proteins, GnRHR1 and GnRHR2 are expressed in the adhesive papillae83. These GnRHRs stimulate cAMP signaling, and the ligand GnRHs for GnRHR1 are also expressed in the papillae83–85. Therefore, GnRHs could be secreted from adhesive papillae as downstream molecules of GABA and stimulate the Gs pathway through GnRHR1. Future studies targeting the regulation of the secretion and reception of GnRH peptides after stimulation of papillae will improve our understanding of the signaling cascade conducting Ciona metamorphosis.

Understanding whether the G-protein signaling relay occurs in cell-autonomous or non-autonomous fashions is crucial for deepening our understanding of Ciona metamorphosis. Each papilla comprises approximately 20 cells, including four papilla neurons (PNs). We showed that activation of Gq and Gs pathways in PNs is sufficient for initiating metamorphosis. This coincides with the previous reports22,23 that the loss of PNs by disrupting the POUIV transcription factor caused the complete arrest of metamorphosis. However, we do not think PNs are the only cells that activate G-proteins, because Ca2+ and cAMP imaging showed upregulation of fluorescence in the entire papillae16,22. A recent study also reports Ca2+ transient in another cell type than PNs34. Other papilla cells are likely responsible for activating G-proteins by directly responding to adhesion and by receiving signal input from adjacent cells, thereby enhancing the signal strength to accumulate a sufficient quantity of cAMP in the PNs to initiate metamorphosis. In this study, technical limitations prevented us from characterizing cells exhibiting increases in Ca2+ and cAMP; we need to address this question in future studies.

Evolutionary implications

Sessile or benthic marine invertebrates lose locomotive activity when they undergo metamorphosis triggered by adhesion to a substratum86. These animals are suspected of having a system to control the adhesion state, allowing them to repeat attachment and detachment before meeting the appropriate conditions for metamorphosis. For example, barnacle larvae “walk” on the substratum before adhering firmly by secreting cement87. Like Ciona, the larvae of sessile/benthic animals may have a system to initiate metamorphosis only when appropriate adhesion is provided, while erasing stimuli from transient and inappropriate adhesion. GABA serves as the metamorphosis inducer of some benthic invertebrates, including mollusks and echinoderms88–90. Moreover, GPCR is implicated as the mediator of settlement and metamorphosis induction in hydrozoans, mollusks, and barnacles88,91,92. In the abalone and barnacle, the dual use of Gq and Gs pathways in metamorphosis has been suggested93,94. Supported by these shared features in the metamorphic mechanisms, our working hypothesis about the initiation of Ciona metamorphosis (Figure 7) will serve as a cue to elucidate how marine benthic invertebrates regulate their metamorphosis.

Materials and Methods

Animals

Ciona intestinalis Type A wild types collected from Onagawa Bay (Miyagi, Japan) and Onahama Bay (Fukushima, Japan) were cultivated in closed colonies by the staff of the National BioResource Project, Japan. They were kept under a constant light condition to prevent gamete release. Eggs and sperm were collected surgically from gonadal ducts, and insemination was carried out in dishes. To prevent larvae from initiating metamorphosis, the tail’s posterior half was manually cut with a scalpel. Removing the tail prevents larvae from swimming efficiently, and these larvae are usually unable to adhere to a substrate. These adhesion-prevented larvae do not metamorphose, since adhesion is required to initiate metamorphosis. Larvae developed from dechorionated eggs were cultured on a 2% agar-coated dish after tail amputation to prevent metamorphosis. Because larvae stick to plastics, tail amputation is insufficient to avoid their adhesion to the culture dish.

Pharmacological treatment

Larvae or larval anterior fragments isolated with a scalpel were administered overnight with 700 μM GABA (Fujifilm Wako #010-02441) or 1 mM theophylline (Sigma-Aldrich #T1633) dissolved in seawater, or 4 μM U0126 (Promega #V1121) dissolved in DMSO.

Plasmids

The open reading frame (ORF) of Kaede was removed from pSPCiPC2Kaede by inverse PCR with PrimeStar GxL DNA polymerase (Takara-bio #R050). The ORFs of Gαq, Gαs, Gαi, and PLCβ1/2/3 were PCR amplified using full-insert cDNAs95. These PCR fragments were fused with the In-Fusion HD Cloning kit (Clontech #639650). The region encoding XY linker43 was deleted from PLCβ1/2/3 cDNA by inverse PCR to create a constitutively active form. The mutations were introduced by inverse PCR to create constitutive active or dominant-negative forms of Gα cDNAs. The introduced mutations are as follows: caGαq, corresponding to Q223L, abolishes GTPase activity42; caGαs, corresponding to Q227L, abolishes GTPase activity47; dnGαi, corresponding to S47C66,67; and dnGαq, corresponding to S54N66. The ORF of bPAC51 was PCR amplified and inserted between the BamHI and EcoRI restriction sites of pSP-Kaede96 to create pSPbPAC. The PCR-amplified PC2 cis-element46 was inserted into the BamHI site of pSPbPAC. The plasmid DNAs were linearized by a restriction enzyme and purified with the Qiaquick Gel Extraction Kit (Qiagen #28706) before microinjection. The ORF of Pink Flamindo52 was PCR amplified and inserted into the EcoRV restriction sites of pBS-HTB97,98 to create pHTBPinkFlamindo. The plasmids pHTBGCaMP822 and pHTBPinkFlamindo were linearized using XhoI for subsequent in vitro synthesis of GCaMP8 and Pink Flamindo mRNA, respectively. GCaMP8 mRNA was synthesized with the MEGAscript T3 kit (Thermo Fisher Scientific #AM1338), the Poly (A) Tailing kit (Thermo Fisher Scientific #AM1350), and the Cap structure analog (New England Biolabs #S1404). Pink Flamindo mRNA was synthesized with the mMESSAGE mMACHINE T3 Transcription kit (Thermo Fisher Scientific #AM1348).

Microinjection

Unfertilized eggs were dechorionated in sterilized seawater containing 1% sodium thioglycolate (Fujifilm Wako #590-11762) and 0.05% actinase E (Kaken Pharmaceutical #650164) as previously described99. The microinjection solution included 2 mg/ml of Fast Green FCF (Fujifilm Wako #061-00031), 0.5-1.0 mM of morpholino oligonucleotide(s) (MOs)100, 5 ng/ul of plasmids, and/or 1 mg/ml of GCaMP8 mRNA for imaging22. For cAMP imaging, 1 mg/ml Pink Flamindo mRNA was dissolved in water without Fast Green, since this chemical emits fluorescence. Microinjected unfertilized eggs were inseminated in a gelatin-coated plastic dish. After the seawater was exchanged to remove excess sperm, the fertilized eggs were cultured at 18 °C or 20 °C overnight until the larval stage. MOs are listed below. Standard control MO, 5’-CCTCTTACCTCAGTTACAATTTATA-3’; GAD ATG MO36, 5’-ACCTCCAAGCCGATTGTTTCTGCAT-3’; GABABR1 ATG MO36, 5’-GCTTACGACTTTACATAACCTTACA-3’; Gαq ATG MO, 5’-GGCATATTTGTGACTATAATGACG-3’; Gαs ATG MO, 5’-AAAGCAACCCATTGGCATTATCGAC-3’; PLCβ1/2/3 splicing MO, 5’-GTGTTACTTACGCTTTCTCTA-3’; PLCβ4 splicing MO, 5’-AACCACCAACCACCAACCTTTTG-3’; IP3R splicing MO, 5’-AATGATGGTTTAAAATTGCCACCTG-3’; Gαi ATG MO, 5’-GTGGAGACTGTGCAACCCATGATTC-3’; dvGai_Chr2 ATG MO, 5’-CCATCTTGAGTAATCCAGGCTTTTA-3’; GIRK channel ATG MO, 5’-TCTGCTGGTTCAGTAATAGACATAG-3’.

Photographing and imaging

Photographs were taken with an AxioImager Z1 and AxioObserver Z1 (Carl Zeiss). The images were treated with AxioVision Rel.4.6 or Zen (Carl Zeiss) and Photoshop 2021 (Adobe) software. Imaging of the Ca2+ transient was carried out according to previous reports16,22. The tails of GCaMP8 mRNA-injected larvae were removed by a scalpel at 15-17 hours after fertilization (hpf). At 30-40 hpf, the larvae were mounted onto a plastic dish under an M165FC stereo microscope (Leica Microsystems), and their adhesive papillae were stimulated by a glass rod using a three-axis coarse manipulator M-152 (Narishige Japan). Immediately after the tip of the rod had contact with the papillae, visible light was shut off and the fluorescence of GCaMP8 was taken with a DFC 310FX digital camera (Leica), Las version 4.12 (Leica), and the movie-capturing function of Windows 10. Images were treated with Photoshop 2021.

Imaging of cAMP was taken by confocal laser scanning microscopy (FV1000, Olympus). The movies were treated with ImageJ (National Institutes of Health). The Pink Flamindo mRNA-injected larvae were immobilized on Poly L lysine-coated glass-bottom dishes at 20-21 hpf. For stimulation of their adhesive papillae with a glass rod, an electric manipulator MM-89 (Narishige) and hydraulic micromanipulator MMO-202ND (Narishige) were used.

RNA sequencing

Larval anterior portions, including adhesive papillae and the other trunk region, were separately isolated by a scalpel around 22-25 hpf at 18°C and collected in Isogen (Nippon Gene #317-02503) soon after isolation. To identify GABA-responsive genes, half of the tail was amputated from larvae around 21-24 hpf, followed by GABA administration for 2 hours before treatment with Isogen. In each experiment, approximately 50 fragments were collected, and two biological replicates were taken. Total RNA was isolated according to the manufacturer’s instructions. Ribosomal RNA was depleted from the total RNA using the NEBNext rRNA Depletion kit (New England Biolabs #E6310), followed by the conversion of the remaining RNA into an Illumina sequencing library using the NEBNext Ultra Directional RNA Library Prep kit (New England Biolabs #E7760). Following library preparation, validation was performed using the Bioanalyzer system (Agilent Technologies) to assess size distribution and concentration. Subsequently, sequencing was carried out on the NextSeq 500 platform (Illumina) employing the paired-end 36-base read option. Reads were mapped on the Ciona intestinalis Type A reference genome (HT model; http://ghost.zool.kyoto-u.ac.jp/default_ht.html) and quantified using CLC Genomic Workbench version 22.0 (Qiagen). RNA-seq data sets were deposited in the NCBI Gene Expression Omnibus (GEO) under accession numbers SAMN40712866 to SAMN40712873, which will appear online upon publication of this paper.

The read counts were normalized by calculating the number of reads per kilobase per million (RPKM) for each transcript in each sample using CLC Genomic Workbench version 22.0 (Qiagen). Eight candidate genes showing a minimum 1.5-fold increase following GABA administration, along with notably higher expression in the papillae in comparison to the trunk (an approximately 4-fold increase in a sample), were chosen for quantitative PCR analysis to assess their response to GABA.

Quantitative PCR

Total RNA was isolated from the anterior tips of larvae treated with the chemical using Isogen (Nippon Gene), following the manufacturer’s instructions. After isopropanol extraction, genomic DNA removal and reverse transcription were carried out using the PrimeScript™ RT reagent kit with gDNA Eraser (Takara Bio #RR047). Quantitative RT-PCR (qRT-PCR) was done using a SYBR Premix Ex TaqII Dimer Eraser (Takara Bio #RR091) and a Thermal Cycler Dice Real Time System III (Takara Bio). The gene encoding elongation factor 1α (EF1α) was used to normalize RNA quantity according to a previous study98. qPCR primers are listed below. KY21.Chr5.240, 5’-TCTTCTCAAAGTTGCACATTCC-3’ and 5’-CAGCAGCAACCAAACGATAAAC-3’; KY21.Chr8.489, 5’-CAATGCAACTTTGACTGCATAC-3’ and 5’-TCCAAACTGCATTCCACATATC-3’; EF1α, 5’-CATGTCACGGACAGCGAAACG-3’ and 5’-CAATGTGTGTTGAGGCATTCCAAG-3’.

Statistical analysis

Differences between conditions were evaluated by Fisher’s exact test. Statistical analyses and most graph visualizations were conducted utilizing R x64.4.1.2 and RStudio software.

Phylogenetic analysis

We performed a tBlastn search101 against the HT gene model of Ciona61 using the amino acid sequences of human PDE, PLCβ, and Ciona Gα proteins as queries, followed by a reciprocal Blast search. We aligned the amino acid sequences from the HDc domain of PDE, the EF-hand_like, PLCXc, PLCYc, and C2 domains of PLCβ, and the whole open reading frame of Gα proteins using the M-COFFEE program102,103. After the removal of unnecessary residues, maximum likelihood trees were constructed using MEGA software version 11104, employing the WAG amino acid substitution matrix105. The trees were assessed with 1,000 bootstrap pseudoreplicates.

eLife Assessment

This important work substantially advances our understanding of the molecular mechanisms underlying the timing of the initiation of metamorphosis of the Ciona ascidian tadpole larva. Through the combination of gene knockdown experiments and fluorescent molecular reporters the authors provide compelling evidence about a crosstalk between different G protein mediated signalling pathways and are able to place different signalling molecules within a signalling network. The work will be of interest to molecular, developmental and marine biologists and to scientists working on animal metamorphosis.

Reviewer #1 (Public Review):

Summary:

In this manuscript, the authors use gene functional analysis, pharmacology and live imaging to develop a proposed model of diverse G protein family signalling that takes place in the papillae during the ascidian Ciona larval adhesion to regulate the timing of initiation of the morphological changes of metamorphosis. Their experiments provide solid evidence that antagonistic G protein signalling regulates cAMP levels in the papillae, which provides a threshold for triggering metamorphosis that is reflective of a larva keeping a strong and sustained level of contact with a substrate for a minimum period of approximately half an hour. The authors discuss their reasoning and address different specific aspects of their proposed timing mechanism to provide a logical flow to the manuscript. The results are nicely linked to

the ecology of Ciona larval settlement and will be of interest to developmental biologists, neurobiologists, molecular biologists, marine biologists as well as provide information relevant to antifouling and aquaculture sectors.

First, they knock down the G proteins Gaq and Gas to show that these genes are important for Ciona larval metamorphosis. They then provide evidence that the Gaq protein acts through a Ca2+ pathway mediated by phospholipase C and inositol triphosphate by showing that inositol phosphate and phospholipase C gene knockdown also inhibits metamorphosis, while overexpression of Gaq or phospholipase C allows larvae to undergo metamorphosis even in the absence of their mechanosensory cue, which is deprived by removing the posterior half of the tail and culturing the larvae on agar-coated dishes. The authors used calcium imaging which is a genetically encoded fluorescent calcium sensor to show that Gq knockdown larvae lack a Ca2+ spike in their papillae after mechanostimulation, confirming that Gaq acts through a Ca2+ pathway. Similarly the authors show that overexpression of Gas also enables larvae to metamorphose in the absence of mechanostimulation, suggesting a role for both Gaq and Gas in this process.

To confirm that Gas acts through cAMP signalling, the authors use pharmacological treatment or overexpression of a photoactivating adenylate cyclase to increase cAMP, and show that this also enables larvae to metamorphose in the absence of mechanostimulation, but only

when their adhesive papillae are still present. Transcriptome data indicate that both Gs and Gq pathway genes are expressed in the adhesive papillae of the Ciona larva. One missing detail seems to be the need for evidence that cAMP is elevated in the papillae directly as a result of Gs activation. The authors use a fluorescent cAMP indicator, Pink Flamindo, to show that cAMP increases in the papillae upon adhesion to a substrate. Complementary to this, larvae that fail to undergo metamorphosis lack a cAMP increase in papillae. However, it is unclear whether the measured larvae that failed to undergo metamorphosis were wildtype or Gas knockdown larvae. If they were Gas knockdown larvae, this could provide evidence that cAMP does act downstream of the Gas activation.

The authors then provide evidence that GABA signalling within the papillae is acting downstream of the G proteins to induce metamorphosis. Transcriptome data shows that the genes for the GABA-producing enzyme, and for GABAb receptors, are both expressed in papillae. Pharmacological experiments show that GABA induces metamorphosis in the absence of mechanosensory cues, but only in larvae that retain their papillae. To show that GABA signalling within the papillae, rather than from the brain of the larva is important, the authors also demonstrate that anterior segments of larvae lacking the brain can also be stimulated to metamorphose by GABA, and show changes in gene expression caused by GABA.

The authors then use a combination of pharmacology and knockdown experiments in the presence or absence of mechanosensory cues to show that Gq/Ca2+ signalling acts upstream of Gs/cAMP signalling. As the elevation of cAMP by pharmacology or photoactivating adenylate cyclase rescued GABA pathway mutant larvae, the Gq and Gs pathways were concluded to be downstream of GABA signaling. However, GABA treatment could still induce Gaq- and Gas-knockdown larvae to metamorphose, suggesting an alternative pathway to metamorphosis, which the authors deduce to be through a third G protein, Gai. They identify an unusual Gai protein that based on transcriptome data is strongly expressed in the papillae. Gai knockdown larvae fail to metamorphose but are rescued by GABA treatment, which can be explained by a potential additional Gai protein being still present (transcriptome evidence suggests this although it is not further confirmed experimentally, for example by hybridization, immunohistochemistry, fluorescent labelling, or knockdown). The authors then use overexpression and knockdown experiments to show that the Gai protein acts through Gβγi complex to activate phospholipase C. Their experiments also indicate a potential for a complementary or compensatory role for Gai and Gaq signalling through Gβγi. By inhibiting the potassium channel GIRK through knockdown, and the MAPK pathway gene MEK1/2 by pharmacology, the authors also establish a role for these in their proposed model of signalling, allowing GABA and cAMP to compensate or interact with each other.

Strengths:

The strength of this paper is the meticulous and extensive experiments, which are carefully designed to be able to precisely target specific genes in the putative signalling pathway to build step by step a complex model that can demonstrate how metamorphosis of the ascidian larva is timed so as to only undergo metamorphosis when strongly attached to a

suitable substrate. The unique possibility of inhibiting mechanosensory-induced metamorphosis by removing some of the tail and smoothing the attachment substrate allows the authors to investigate potential effects on both activation and inhibition of metamorphosis, and to confirm that specific signalling pathways are clearly downstream of the initial

mechanosensory stimulation. The study is also clear about which aspects of the model still remain unknown, such as which ligands and receptors may be responsible for the binding and activation of Gaq and Gas. Experiments testing metamorphosis of just the anterior region of the larvae nicely demonstrates the need for signalling in the region of the papillae, as do experiments where the papillae are removed, which then block metamorphosis in treatments that would otherwise stimulate it. The final model is a nice end point and makes a clear summary of how the extensive experiments all fit together into a cohesive potential signalling network, which can be built upon in the future.

Weaknesses:

The paper has few weaknesses, however the main difficulty it poses is that due to the sheer number of precise experiments carried out and the complexity of the interwoven signalling pathways, it quickly becomes very difficult to follow exactly what is going on when and why or to keep track of the story as it develops. To improve this, an initial section in the results could be included showing a summary of the known G proteins in Ciona, their types and potential downstream signalling or upstream receptors, where known, and their expression levels in papillae. This could be in the form of a table and/or include the phylogenetic tree from the supplementary data. This would help clarify why the study first focuses on Gaq and Gas, and only later looks at Gai. This could be supplemented by a schematic workflow giving an overview of the experimental process of the study. A second minor weakness (understandable as the focus of the study is metamorphosis induced by mechanosensory stimulation) is that the study does not take into account any potential role for other types of sensory modalities (light, chemicals) that may also feed into the regulation of Ciona larval metamorphosis. This aspect would be interesting to discuss in light of the recent paper suggesting that some sensory cells in the Ciona adhesive papillae are polymodal and detect both chemicals and mechanical stimuli (Hoyer et al. 2024 Current Biology 34(6): 1168 - 1182).

Reviewer #2 (Public Review):

Summary:

This work aims to characterize the neural signaling cascade underlying the initiation of metamorphosis in Ciona larvae. Combining gene-specific functional analyses, pharmacological experiments, and live imaging approaches, the authors identify the molecular players downstream of GABA to initiate Ciona metamorphosis. The results of this study may serve as a useful framework for future research on animal metamorphosis.

Strengths:

The authors did a great job in connecting their experiments with previous findings on Ciona metamorphosis. Taking advantage of the Ciona model system, they meticulously conducted genetic manipulation and pharmacological experiments to test the epistatic relationships among the signaling players controlling the initiation of Ciona metamorphosis.

Weaknesses:

The causal relationship between cAMP accumulation and the initiation of metamorphosis was not clearly demonstrated by the life-imaging observation with the fluorescent cAMP indicator (Pink Flamindo). It is a pity that this experiment was only conducted using normal larvae to compare those who underwent metamorphosis versus those who failed to initiate metamorphosis. This approach should be applied to some of the genetic manipulation and pharmacological experiments, to strengthen their main thesis on the "cAMP timer" mechanism.

On several occasions, the interpretation of the results seems to be imprecise and may lead to misunderstanding. This should be improved by rewriting the descriptions of those results and carefully comparing the differences in results from various treatments and experiments.

Author response:

The following is the authors’ response to the original reviews

Thank you for your valuable comments, which helped us improve our manuscript. We will make the following modifications in the revised manuscript:

(1) In the first paragraph of the Result section, we will provide a summary of trimeric G proteins in Ciona and explain how we focused on Gαs and Gαq in the initial phase of this study.

We added a summary of trimeric G proteins in Ciona in the initial part of the Results section (page 6, line 23 to page 8, line 5). In this summary, we added the following sentence explaining the reason we focused on Gas and Gaq in the initial phase of this study: "Among them, we prioritized examining the Gα proteins having an excitatory function (Gαq and Gαs) rather than inhibitory roles since previous studies suggested that excitatory events like Ca2+ transient and neuropeptide secretion occur when Ciona metamorphose."

(2) As the reviewer 1 suggests, the polymodal roles of papilla neurons are interesting. Although we could not address this through functional analyses in this study, we will add a discussion regarding this aspect. The sentences will be something like the following:

“The recent study (Hoyer et al., 2024) provided several lines of evidence suggesting that PSNs can serve as the sensors of several chemicals in addition to the mechanical stimuli. This finding and our model could be mutually related because these chemicals could modify Ca2+ and cAMP production. The use of G protein signaling allows Ciona to reflect various environmental stimuli to initiate metamorphosis in the appropriate situation, both mechanically and chemically.”

We added a discussion related to the recent publication by Hoyer and colleagues on page 23, lines 13-18: " A recent study[19] provided several lines of evidence suggesting that PNs can serve as the sensors of several chemicals in addition to mechanical stimuli. This finding and our model could be mutually related because these chemicals could modify Ca2+ and cAMP production. G protein signaling allows Ciona to reflect various environmental stimuli to initiate metamorphosis either mechanically or chemically according to the situation."

(3) As both reviewers suggested, imaging cAMP on the backgrounds of some G protein knockdowns is essential, and we will conduct the experiments.

We added the data on cAMP imaging in Gas, Gaq, and dvGai_Chr2 knockdown larvae in Supplementary Figure S4C-D and Figure 6E.

(4) We carefully modify the text throughout the manuscript so that the descriptions suitably reflect the results.

We modified the descriptions of experimental results so that the text reflects the results more precisely.

Reviewer #1:

Pg1 - need to add an additional '6' to the author list to clarify which two or more authors contributed equally.

We added a 6 as suggested. Thank you for pointing this out.

Pg3 - note that larval adhesive organ applies to not all benthic adults, but to benthic sessile adults this makes it sound like the adhesive organ can trigger metamorphosis but has that been shown? In Ciona or others? Need to specify the role of cells secreting adhesive, vs sensory cells that trigger metamorphosis?

We divided the corresponding sentence into two to clearly state that adhesion and triggering metamorphosis are related but could be different events. Moreover, we modified the sentence to state that physical contact is one example of a cue triggering metamorphosis. We then added another example of a factor triggering metamorphosis—i.e., chemicals from the organisms surrounding the adherence site (page 3, lines 16-20 of the revised version):

"Many marine invertebrates exhibit a benthic lifestyle at the adult stage[4]. Their planktonic larvae have an adhesive organ that secretes adhesives and adheres to a substratum. The cues associated with the adhesion, such as the physical contact with the substratum and a chemical from organisms surrounding the adherence site, can trigger their metamorphosis."

Pg 4 - although mechanosensation is the focus here, could there also be chemoreception/chemoreceptors involved in Ciona metamorphosis? For example, Hoyer et al. 2024 (Current Biology 34(6):1168-1182) concluded that some palp sensory neurons were multimodal and could be both chemo- and mechano-sensory.

We added statements about this recent finding in the Introduction and Discussion sections. In the Introduction (page 4, lines 16-18), however, we also stated that a mechanical stimulus can trigger metamorphosis in the lab without the need to supply these chemicals. This is to emphasize that the mechanical stimulus is the focus of this study. In the Discussion, we added a statement that G-protein signaling could also be used to receive the chemical stimuli (page 23, lines 13-18).

Pg 6 - Before starting functional characterizations, it would be useful to give an overview (table?) of the G proteins found in papillae, and what receptor they are suspected of binding to, or if this is completely unknown, and which downstream pathways they likely activate. That is, to show some results about which G proteins are found in Ciona, and which are found in papillae. In this way, it will make more sense for readers when the Gai is suddenly introduced later, following the sections of Gaq and Gas.

Thank you for your idea to improve the readability of this manuscript. In the initial part of the Results section (page 6, line 22 to page 8, line 5), we added descriptions of the repertoire of trimeric G-proteins in Ciona, including phylogenetic analyses, and expression in the papillae based on RNA-seq data, followed by the reason why we initially focused on Gaq and Gas. The data are displayed in Supplementary Figure S1. The phylogenetic analyses were modified from those shown in Supplementary Figure S5 of the previous version. We also added the general downstream activities of Gas, Gai and Gaq in the Introduction section (page 6, lines 10-12). Considering the contents, the general function of Ga12/13 was stated in the Results section (page 8, lines 2-3).

We did not add the information about their partner receptors in this early section. This is because there are many candidates, and we could not pick some of them. Instead, we described our current suppositions about their possible partners in the Discussion (page 23, line 22 to page 24, line 19). However, we suspect that there are more candidates, and we wish to promote unbiased research in the future.

Pg 9 - would be good to know the timing of this PF fluorescence increase and the timing of stimulation in the text here, relevant to the 30-min gap before metamorphosis initiation

We added the start times for the cAMP reduction and re-upregulation in the following sentence (page 11, lines 17-18): "The cAMP reduction and increase respectively started at 35 seconds and 4 min 40 seconds after stimulation on average."

Pg 28 - Phylogenetic analysis: Given that the results may be of interest to metamorphosis in other marine invertebrates as discussed in the last paragraph of the paper, it would be useful to include G proteins from these other animal phyla where available in the phylogenetic tree. Similarly, in Figure S5A it would be useful to highlight further all the different Ciona G proteins, and the different protein families, through the use of additional colour/labelling (regardless of whether this remains Fig S5A, or becomes part of the main figures)

We drew a phylogenetic tree of G-proteins including those in some sessile and benthic animals (barnacle, sea anemone, hydra, sponge, sea urchin and shell). However, we decided not to add the tree in the revised version because, unfortunately, the bootstrap values of many branches were not high enough to have confidence in the results. We hope you understand our decision. Ciona divergent G-proteins are likely to be specific to Ciona.

According to your comment, we highlighted all Ciona G alpha proteins in red in Figure S5A, which is now Figure S1A in the revised version.

Figure 3E and Figure S3 - is the data shown as an average of all larvae measured (n=5 and n=4) or is it data from one representative larva out of the 4-5 measured? This needs clarification.

The original graphs in Figure 3E and Figure S3 are typical examples. We added the graphs summarizing data of all larvae in each experimental condition in Supplementary Figure S4 (corresponding to Supplementary Figure S3 of the original version). Figure 3E remains as a typical example of the result of a single larva to explain our data analysis in detail.

Experimental suggestion - As mentioned above, one missing detail seems to be the need for evidence that cAMP is elevated in the papillae directly as a result of Gs activation- this could be shown with measurement of cAMP via PF in Gs knockdown larvae that are mechanically stimulated compared to wildtype stimulated and non-stimulated?

Thank you for your suggestion. The experiments are indeed important. We added the data of Pink Flamindo imaging in the Gas, Gaq and dvGai_Chr2 knockdown conditions. The results of Gas and Gaq knockdowns are described in page 11, line 24 to page 12, line 5, and are displayed in Supplementary Figure S4C-D. The result of dvGai_Chr2 knockdown is given on page 16, lines 20-22 and shown in Figure 6E.

In order to insert the data of cAMP imaging of dvGai_Chr2 knockdown larvae, we transferred some panels of Figure 6 to Supplementary Figure S6. In addition, the knockdown data of dvGαi_Chr4 and double knockdowns of Gai genes are also included in Supplementary Figure S6.

Reviewer #2:

Page 6, line 3-4 in the first paragraph of the "Results"; the authors state "Neither morphant showed any signature of metamorphosis even though both were allowed to adhere to the base of culture dishes...". However, judging from Fig. 1E, "the percentage of metamorphosis initiation" (indicated by the initiation of tail regression) in Gαq morphans is not close to 0 (average about 40%), thus I am not convinced this observation can be described as "Neither morphant showed any signature of metamorphosis..." in this sentence.

Thank you for your suggestion. In writing the original text, we oversimplified some of the descriptions when trying to improve the readability. We agree this resulted in imprecision in places. We have revised all these passages in our revision. In this particular case, we softened the overly emphatic statement to better reflect the results, changing “... any signature of metamorphosis...” to “... reduced rate of metamorphosis initiation...” In addition, we stated that the effect of Gαq MO was weaker than that of Gαs MO on page 8, lines 10-12. The weaker effect of Gaq MO was due to the redundant role of the Gi pathway, which is shown on page 17, lines 10-17, and in Figure 6G-H.

Similarly, in the next paragraph describing the knockdown of PLCβ1/2/3, PLCβ4, and IP3R genes, the authors appear to neglect there is a weaker effect of the PLCβ4 MO, and simply described the results as "The knockdown larvae of these three genes failed to start metamorphosis". Based on Fig. 1H, about 30% of the PLCβ4 MO-injected animals still initiated tail regeneration. This difference may have some biological meanings and thus should be described more precisely.

We added the following sentence on page 8, lines 18-19 of the revised version: “The effect of PLCβ4 MO was weaker than those of the other MOs, suggesting that this PLC plays an auxiliary role.”

Page 7, second paragraph, on the description of GCaMP8 fluorescence and also at the end of Fig. 1O legend, the citation to "Figure S1" is confusing; Fig. S1 is the phylogenetic tree of PLCβ proteins. Is there additional data regarding this Gαq MO plus GCaMP8 mRNA injection experiment?

Figure S1 of the original version corresponds to Figure S2 of the revised version. To avoid confusion, we deleted this citation from the legend of Figure 1O. By this modification, the sentence stating the repertoire of PLCb and IP3R in Ciona (page 8, lines 15-16) is the only sentence citing Figure S2 in the revised version.

Page 8, first sentence; The purpose of theophylline treatment is not to prevent larvae from adhesion, thus I would suggest modifying this sentence to: "We treated wild-type larvae with theophylline after tail amputation, and we observed that most theophylline-treated larvae completed tail regression without adhesion (Figure 2D-F)".

We modified the sentence according to your comment. Thank you for your suggestion.

Page 9, second paragraph; judging from the data presented in Fig. 3C, I think this description: "when papillae were removed from larvae, theophylline failed to induce metamorphosis" is not accurate, because about ~30% of the Papilla cut +Theophylline-treated larvae still initiated their tail regression. This needs to be explained clearly.

We modified the sentence (page 11, lines 2-3) as follows: “...the average rate of metamorphosis induction by theophylline was reduced from 100% to 30%...”

Similarly in the next few sentences regarding the results presented in Fig, 3D, the effects of overexpressing those genes are not uniform. While amputation of papillae in larvae overexpressing caPLCβ1/2/3 could inhibit metamorphosis almost completely, papilla cut seems to have a weaker effect on caGαq, caGαs, and bPAC-overexpressing larvae.

We added a description explaining that caPLCβ1/2/3 was the most sensitive to papilla amputation, and the possibility that PLCβ1/2/3 works specifically in the papillae (page 11, lines 9-11): “Among these experiments, caPLCβ1/2/3 overexpression was the most sensitive to papilla amputation, suggesting that PLCβ1/2/3 acts specifically in the papillae during metamorphosis.”

Page 9, the paragraph on using the fluorescent cAMP indicator; there is a discrepancy between the described developmental time when the authors conducted this experiment and the metamorphosis competent timing (after 24hpf) described on page 7. On page 26, the authors describe "The Pink Flamindo mRNA-injected larvae were immobilized on Poly L lysine-coated glass bottom dishes at 20-21 hpf...". Did the authors start stimulating the larvae to observe the fluorescent signal soon after immobilization, or wait several hours until the larvae passed 24hpf and then conduct the experiment?

The latter is the case. The immobilized larvae were kept until they acquired the competence for metamorphosis and then stimulation/recording was carried out. This point is described in the Materials and Methods section of the revised version (page 29, lines 16-18):

"The Pink Flamindo mRNA-injected larvae were immobilized on Poly L lysine-coated glass-bottom dishes at 20-21 hpf, and stimulated their adhesive papillae around 25 hpf."

Page 10, the description "...Gαq morphants initiated metamorphosis when caGαs was overexpressed in the nervous system (Figure 4F)". It should be noted that the result is only a partial rescue. To be precise, this description needs to be modified.

We changed the sentence to reflect the results more precisely (page 14, lines 2-3): “Moreover, caGαs overexpression in the nervous system significantly, although not perfectly, ameliorated the effect of Gαq MO (Figure 4F).”

Page 12-13, This description and the figure 5E presented is a bit confusing to me. The figure legend for 5E: "GABA is necessary for Ca2+ transient in the adhesive papillae (arrow)" But the arrow in this image points to a place with no fluorescent signal, and on the upper corner it labeled as "29% (n=17)". Does that mean the proportion of "no Ca2+ increase after stimulation" was 29% among the 17 samples examined? Or actually, is the other way around that 81% of the examined larvae did not show Ca2+ signal increase after stimulation?

The latter is the case. We added a caption explaining this clearly in the Figure legend: “The percentage and number exhibit the rate of animals showing Ca2+ transient in the papillae.”

Page 13, second paragraph; I do not agree with the overly simplified description that "GABA significantly ameliorated the metamorphosis-failed phenocopies of Gαq, PLCβ, and Gαs morphants". As shown in Fig. 5F-H, adding GABA exerts different levels of partial rescue effect on each morphant, and thus should be described clearly.

When the outliers are neglected, the effect of GABA is most evident in Gαs knockdowns. This suggests that the target(s) of GABA signaling is more likely to be Gq pathway components. We added the following sentence to the revised version (page 15, lines 14-16):

“Among the three morphants, GABA exhibited the most effective rescues in Gαs knockdowns than Gαq and PLCβ.”

In addition, we think this sentence establishes a more logical connection with the sentence that follows it: “These results could be explained by assuming enhancement of the Gq pathway by GABA through PLCβ and another GABA-mediated metamorphic pathway bypassing Gq components.” Thank you for your suggestion.

The section "Contribution of Gi to metamorphosis" confirmed the possibility that GABA signaling targets Gq pathway components.

Page 13, the first paragraph on "Contribution of Gi to metamorphosis"; the description that "The knockdown of this gene (Gαi) exhibited a significantly reduced rate of metamorphosis;..." is misleading. I would suggest modifying the entire sentence as "The knockdown of this gene (Gαi) exhibited a moderate (although statistically significant) reduction of metamorphosis rate, suggesting the presence of another Gαi regulating metamorphosis".

Thank you for your suggestion. We modified the sentence (page 16, lines 2-4 in the revised version) as recommended. We believe the description is much improved.

Page 20, the last sentence about Ciona papilla neurons expressing transcription factor Islet; the authors seem to attempt to make some comparison with the vertebrate pancreatic beta cells in this paragraph, but the comparison and the argument are not fully developed in this current format.

To deepen this discussion, we added the following sentence (page 23, lines 10-12): “The atypical secretion of GABA might depend on the transcription factor like Islet shared between Ciona papilla neurons and vertebrate beta cells.”

However, we would like to limit the depth of our discussion on this point, as we hope to expand on it further in future studies.

Other suggestions:

Page 3, second paragraph: as they become unable to "move" after metamorphosis -> "relocate"

We corrected the word as suggested.

Page 4, second paragraph: In the first sentence, the author states the current understanding of chordate phylogeny and cites Delsuc et al. 2006 Nature paper at the end of this sentence. However, in this paper cephalochordates were erroneously grouped with echinoderms, and thus chordates did not form a monophyletic clade. A later paper by Bourlat et al, (Nature 444:85-88, 2006) corrected this problem, and subsequently Dulsuc et al. also published another paper (genesis, 46:592-604, 2008) with broader sampling to overcome this problem. These later publications need to be included for the sake of correctness.

We added this reference.

Page 14, regarding the redundant function of the typical Gαi protein in the papillae; the authors may try double KD of Gαi and dvGαi_Chr2 in their experimental system to test this idea.

We carried out double knockdown of typical Gai and dvGαi_Chr2. However, we could not address their redundant role sufficiently because most of the double knockdown larvae exhibited severe shape malformation.

dvGαi_Chr4 is also expressed in the papillae. We carried out knockdown of this gene, to find that the knockdown resulted in very minor but statistically significant reduction of the metamorphosis rate, suggesting that this Gai also plays a supportive role in metamorphosis. We also carried out double knockdown of dvGαi_Chr2 and dvGαi_Chr4. The double KD larvae exhibited responsiveness to GABA, probably because of the presence of typical Gai.

These results are described on page 16, lines 2-18, and the data are shown in Supplementary Figure S6A-D of the revised version.

Responses to the Reviewing editor's comments:

"Larvae of the ascidian Ciona initiate metamorphosis tens of minutes after adhesion to a substratum via its adhesive organ." - Larvae is plural so change to 'via their adhesive organ'

The sentence was corrected as suggested.

"Metamorphosis is a widespread feature of animal development that allows them" - revise the sentence, e.g. "Metamorphosis is a widespread feature of development that allows animals"

The sentence was corrected as suggested.

"GABA synthase (GAD)" GAD is not called GABA synthase but glutamate decarboxylase - clarify, e.g. encoding the enzyme synthesizing GABA called glutamate decarboxylase (GAD)

This part was corrected exactly as suggested. Thank you.

"IP3 is received by its receptor on the endoplasmic reticulum (ER) and releases calcium ion (Ca2+ )" revise to "IP3 is received by its receptor on the endoplasmic reticulum (ER) that releases calcium ion (Ca2+ )"

The sentence was corrected as suggested.

"Moreover, GPCR is implicated as the mediator of settlement" - GPCRs are implicated

This sentence was modified as suggested.

Version 2

Abstract

Abstract

Larvae of the ascidian Ciona initiate metamorphosis tens of minutes after adhesion to a substratum via their adhesive organ. The gap between adhesion and metamorphosis initiation is suggested to ensure the rigidity of adhesion, allowing Ciona to maintain settlement after losing locomotive activity through metamorphosis. The mechanism producing the gap is unknown. Here, by combining gene functional analyses, pharmacological analyses, and live imaging, we propose that the gap represents the time required for sufficient cAMP accumulation to trigger metamorphosis. Not only the Gs pathway but also the Gi and Gq pathways are involved in the initiation of metamorphosis in the downstream signaling cascade of the neurotransmitter GABA, the known initiator of Ciona metamorphosis. The mutual crosstalk of stimulatory and inhibitory G-proteins functions as the accelerator and brake for cAMP production, ensuring the faithful initiation of metamorphosis at an appropriate time and in the right situation.

Introduction

Metamorphosis is a widespread feature of development that allows animals to have different functions between larval and adult stages1. As larvae become adults, the shape and various characteristics of their physiology, gene expression, behavior, and lifestyle change. With these changes, animals can focus on a limited number of biological activities at each stage to increase feeding, growth, dispersal, and reproduction efficiencies. This biphasic mode of life is a widely conserved and ancient trait of animals, as evidenced by its presence in groups that maintain primitive states2, indicating that this long-conserved feature has contributed to animals’ flourishment. Therefore, to understand animal evolution, it is essential to elucidate the mechanisms underlying metamorphosis.

A key characteristic of metamorphosis is that both internal and external conditions determine its initiation. Sufficiently matured (or metamorphically competent3) larvae start metamorphosis only when they meet an appropriate external condition to be adults. Various organic and inorganic external stimuli suited to the lifestyle of adulthood trigger metamorphosis. Many marine invertebrates exhibit a benthic lifestyle at the adult stage4. Their planktonic larvae have an adhesive organ that secretes adhesives and adheres to a substratum. The cues associated with the adhesion, such as the physical contact with the substratum and a chemical from organisms surrounding the adherence site, can trigger their metamorphosis. Upon starting metamorphosis, their larvae lose locomotive organs and transition into benthic adult forms. Determining the timing of metamorphosis is important for benthic animals, particularly sessile ones, to ensure their future survival and reproduction, as they become unable to relocate after metamorphosis. Therefore, sessile animals are suspected to have elaborate mechanisms to start metamorphosis only when a firm adhesion is achieved at an appropriate place. Although several reports describe this kind of phenomenon5–7, the mechanism by which the external condition is turned into an internal mechanism to trigger metamorphosis at the right timing remains elusive.

Ascidians are marine invertebrate chordates that are the closest living relatives to vertebrates8–11. The chordate features of ascidians are best represented by their tadpole larval shape. Like vertebrate tadpoles, ascidian larvae swim by beating their tail through neuromuscular activities12,13. However, ascidians lose the tadpole shape during metamorphosis, and they exhibit a sessile lifestyle at the adult stage14,15. Ascidian metamorphosis comprises several key events, such as tail regression, body axis rotation, and adult organ growth16 (Figure 1A).

Gq and Gs pathways are required for Ciona metamorphosis.

(A) Schematic illustration of Ciona metamorphosis.

(B) A larva knocked down with Gαq using the antisense morpholino oligonucleotide (MO). Metamorphosis did not initiate at 2 days post-fertilization (2 dpf).

(C) A Gαs knockdown larva.

(D) A control animal injected with the standard (STD) MO. Metamorphosis initiated, as indicated by the completion of tail regression.

(E) Effect of Gαq knockdown on the percentage of metamorphosis initiation (indicated by the initiation of tail regression), shown as box-and-whisker plots. Dots indicate experiment replicates. ***, p<0.001 (Fisher’s exact test). n, number of examined larvae in total.

(F) Effect of Gαs knockdown.

(G-I) Effects of PLCβ1/2/3, PLCβ4, and IP3R knockdowns.

(J-M) Results of the experiments where adhesion was prevented by tail amputation and laying larvae on agar-coated plates.

(J) A control larva overexpressing the Kaede reporter in the entire nervous system using the PC2 cis element.

(K) A larva overexpressing constitutive active (ca)PLCβ1/2/3.

(L) Effect of caGαq overexpression.

(M) Effect of caPLCβ1/2/3 overexpression.

(N-O) Live imaging of Ca2+ transient, monitored by injecting GCaMP8 mRNA.

(N) Snapshot of the control larva injected with GCaMP8 mRNA. An increase in the fluorescence intensity in the adhesive papillae (arrow) was observed. Rod, the position of the glass rod used to stimulate papillae. The percentage and number exhibit the rate of animals showing Ca2+ transient in the papillae.

(O) A larva injected with Gαq MO plus GCaMP8 mRNA. An increase in the fluorescence intensity in the papillae (arrow) was not observed even though the brightness of the whole body was raised.

Adhesion to a substratum triggers ascidian metamorphosis16. Ascidian larvae have adhesive organs called adhesive papillae, usually triangular protrusions at the anterior end. Adhesive papillae secrete a lectin-type mucus substance that is suspected to aid in binding to the substratum17. Moreover, adhesive papillae serve as mechanical sensors18. Although cells in the papillae are polymodal, sensing both mechanical and chemical stimuli19, the mechanical stimulus can trigger metamorphosis without such a chemical18,20 under laboratory conditions. The papilla is a neuronal organ that includes epidermal neurons positive for vesicular glutamate transporter (VGLUT) and GABA, which innervate axons toward the sensory vesicle (larval brain)21–24. The papilla neurons are responsible for initiating metamorphosis18,25–27. Moreover, our recent study suggests the involvement of a mechanoreceptor channel (the TRP channel) in the responsiveness to mechanical stimuli26. However, a simple physical adhesion is not sufficient to trigger metamorphosis. In our study using the model ascidian Ciona intestinalis Type A (it has been proposed that this species be renamed C. robusta28–30, and hereafter we call it Ciona), continuous adhesion for about 30 min is necessary before starting metamorphosis31. When larvae detach before reaching the critical period, they must adhere for 30 min again for metamorphosis. Therefore, it is thought that ascidian larvae somehow sense the duration of adhesion and initiate metamorphosis only when adhesion is firm enough to be maintained for the required period. Ciona larvae continue tail beating during adhesion to push their body to the substratum. We showed that the strength of force generated by this swimming activity influences the timing of metamorphosis initiation26. In this phenomenon, abolishment of swimming activity elongates the time from settlement until metamorphosis is initiated. The mechanisms for measuring the duration of adhesion and the strength of the force generated by adhesion are unknown.

To elucidate the mechanisms triggering ascidian metamorphosis, the signaling pathways must be characterized. Many studies have discovered signaling molecules as possible inducers of ascidian metamorphosis19,32–38. These molecules include neurotransmitters, suggesting that transmission of excitation in the nervous system, starting from the adhesive papillae, is crucial for metamorphosis. Recently, our group reported that the inhibitory neurotransmitter GABA plays a pivotal role in initiating Ciona metamorphosis39. Knockout and knockdown of the genes encoding the enzyme synthesizing GABA called glutamate decarboxylase (GAD), vesicular inhibitory amino acid transporter (VIAAT or VGAT), and metabotropic GABA receptor (GABABR), resulted in the perturbation of metamorphosis. Moreover, GABA administration induces metamorphosis without adhesion. Our previous studies demonstrated that GABA induces the secretion of gonadotropin-releasing hormone (GnRH)39,40; however, it remains unknown how the inhibitory neurotransmitter activates neuronal functions for initiating metamorphosis.

In this study, we addressed the characterization of the downstream cascade stimulated by GABA. Because GABABR is a G-protein-coupled receptor (GPCR)41, we searched for trimeric G-proteins necessary for metamorphosis. These G-proteins are heterotrimers of an α β and γ subunit42. Upon ligand binding, GPCR exchanges GDP of the α subunit to GTP, then the GTP-bound α subunit and the βγ complex are released from the GPCR. Although both the GTP-bound α subunit and the βγ complex have activities, the α subunit mainly determines the reaction specific to the G-protein type. Gαs and Gαi respectively activate and inhibit cyclic adenosine monophosphate (cAMP) synthesis. In contrast, Gαq increases intracellular Ca2+ ion concentration by promoting its secretion from the endoplasmic reticulum. We found that three G-proteins are activated in the downstream GABA cascade, resulting in cAMP elevation. Because the signaling includes stimulating and inhibiting cAMP synthesis, Ciona initiates metamorphosis only when a sufficient quantity of cAMP is accumulated due to sustained adhesion. Our results revealed the ingenious mechanism that permits Ciona to start metamorphosis only when achieving an appropriate adhesion firm enough to relinquish swimming ability and commence sessile adult life.

Results

Characterization of G-proteins necessary for metamorphosis initiation

A previous study43 identified ten genes encoding Gα proteins from the Ciona genome, and we found nine gene models corresponding to these proteins in the latest genome assembly44 (Table S1). Gα proteins are classified into four major groups (Gαs, Gαi, Gαq, and Gα12/13), which have distinct functional properties45. Ciona has one gene encoding an unambiguous ortholog in each of the four groups (Figure S1A). Moreover, a previous study43 suggested the presence of another Gαq protein in the Ciona genome. Our phylogenetic tree suggested that Gαq (Gαq_Chr11) is orthologous to human GNA15 (Figure S1A).

The remaining four Ciona genes (their gene models are KY21.Chr2.875, KY21.Chr4.943, KY21.Chr9.455, and KY21.Chr8.580) encode divergent Gα proteins. The Gα protein encoded by KY21.Chr8.580 has truncation at its N-terminal part, and we omitted this protein from the phylogenetic analysis. Our phylogenetic analyses showed that the remaining divergent Gα proteins have a strong affinity to the Gαi/o family proteins (Figure S1A). Moreover, the amino acid residues at the C-terminal end of these proteins are more similar to human Gαi than the other Gα protein families (Figure S1B). The C-terminal residues, particularly glycine at position 3, are essential for determining the Gi partner of GPCR46. We tentatively name these proteins dvGαi_Chr2, dvGαi_Chr4, and dvGαi_Chr9. Gαi/o family proteins have inhibitory activity that suppresses cAMP production47.

To verify the possible contribution of G proteins to the initiation of metamorphosis, we quantified the expression level of the Gα genes in the adhesive papillae using RNA-seq surrounding this region (Figure S1C and Table S1). Typical Gαq and three Gαi genes exhibited more abundant expression in the papillae, followed by Gαs and Gα12/13. Therefore, these Gα proteins were chosen as candidates for the protein(s) regulating metamorphosis initiation. Among them, we prioritized examining the Gα proteins having an excitatory function (Gαq and Gαs) rather than an inhibitory role, since previous studies suggested that excitatory events such as Ca2+ transient and neuropeptide secretion occur when Ciona metamorphose18,39. The primary function of Gα12/13 is the activation of Rho-mediated actin dynamics. Although this regulation is essential for tail regression48, this event occurs after metamorphosis initiation. Therefore, we omitted this gene from further analysis in this study.

We knocked down the Gαq and Gαs genes using antisense morpholino oligonucleotides (MOs). We found that both these genes are necessary for metamorphosis. The morphants of these genes showed a reduced rate of metamorphosis initiation even though they were allowed to adhere to the base of culture dishes, which was sufficient for control larvae to initiate metamorphosis (Figure 1A-F). Gαs MO exhibited almost complete perturbation of metamorphosis, while the effect of Gαq MO was not strong (Figure 1E and F).

Gαq activates phospholipase C beta (PLCβ) to produce inositol triphosphate (IP3)49. IP3 is received by its receptor on the endoplasmic reticulum (ER) that releases calcium ion (Ca2+). The Ciona genome encodes two PLCβ (PLCβ1/2/3, PLCβ4) and one IP3 receptor (IP3R) (Figure S2 and Table S1). We knocked down the PLCβ1/2/3, PLCβ4, and IP3R genes. The knockdown larvae of these three genes failed to start metamorphosis (Figure 1G-I). The effect of PLCβ4 MO was weaker than those of the other MOs, suggesting that this PLC plays an auxiliary role. These results suggest that Gαq initiates metamorphosis by the conventional Ca2+ pathway mediated by PLCβ and IP3/IP3R. To confirm this, we overexpressed constitutively active forms of Gαq (caGαq)50 and of caPLCβ1/2/351 throughout the entire nervous system with the cis element of the gene encoding prohormone convertase 2 (PC2)52–54. When the microinjected animals reached the larval stage, they were cultured on agar-coated dishes after tail amputation to prevent adhesion. In this condition, the control larvae rarely started metamorphosis because of the absence of adhesion (Figure 1J). In contrast, caGαq- or caPLCβ1/2/3-overexpressed larvae initiated metamorphosis without adhesion (Figure 1K-M).

Adhesive papillae exhibit a Ca2+ increase soon after sensing an adhesive stimulus18,26. We examined whether this Ca2+ transient is dependent on the Gq pathway. Compared to controls, significantly fewer larvae injected with Gαq MO plus GCaMP8 mRNA exhibited GCaMP8 fluorescence elevation in the adhesive papilla upon stimulation (Figure 1N, O). This result suggests that the Gq pathway is activated upon adhesion to cause a Ca2+ transient in the adhesive papilla.

The Gs pathway initiates metamorphosis by activating cAMP synthesis

The involvement of Gs in metamorphosis was confirmed by the overexpression of a constitutively active form of Gαs (caGαs)55 in the nervous system. This overexpression resulted in the enhanced initiation of metamorphosis without adhesion (Figure 2A-C). The Gs pathway activates adenylate cyclase (AC) to produce cAMP56. We previously reported that cAMP can induce metamorphosis40. Because externally added cAMP is not a strong inducer of metamorphosis, we attempted to confirm this hypothesis through another experiment. Theophylline increases cAMP by inhibiting the cAMP-degrading enzyme phosphodiesterase (PDE)57. We treated wild-type larvae with theophylline after tail amputation, and we observed that most theophylline-treated larvae completed tail regression without adhesion (Figure 2D-F). Theophylline has several target proteins in addition to PDE58. To further confirm that cAMP is responsible for the initiation of metamorphosis, we overexpressed photo-activating AC (bPAC)59 in the nervous system.

cAMP initiates Ciona metamorphosis.

(A) An adhesion-prevented control (uninjected) larva at 2 dpf. Metamorphosis did not initiate.

(B) An animal with adhesion prevention and overexpression of caGαs in the entire nervous system. Metamorphosis initiated.

(C) Effect of caGαs overexpression.

(D) A control (uninjected) larva.

(E) A theophylline-treated animal. Metamorphosis initiated.

(F) Effect of theophylline treatment.

(G) Effect of bPAC overexpression in the nervous system.

(H) Effect of bPAC overexpression over time. In this experiment, adhesion was permitted. Control groups (overexpressed with mCherry) exhibited a slower initiation of metamorphosis than bPAC-overexpressed groups.

The bPAC-overexpressed larvae regressed their tails without adhesion, suggesting that a cAMP increase triggers metamorphosis (Figure 2G).

Using bPAC, we addressed whether enhanced cAMP production facilitates the initiation of metamorphosis. At 24 hours post-fertilization (hpf), only a few animals in both the bPAC-overexpressed and control groups initiated metamorphosis because 24 hpf is somewhat too early for Ciona larvae to be metamorphically competent31 (Figure 2H). Even in this condition, the bPAC-overexpressed group exhibited a statistically higher rate of metamorphosis-initiated larvae. The proportion of animals initiating metamorphosis increased over successive time points, with the bPAC-overexpressed group consistently showing a statistically higher rate of metamorphosis compared to controls. These results support the idea that cAMP plays a role as a timer for Ciona metamorphosis; accumulation of this molecule to exceed a threshold could trigger metamorphosis. Because many of the bPAC-overexpressed larvae did not initiate metamorphosis at 24 hpf, similar to control larvae, cAMP accumulation does not alter the timing of acquiring metamorphic competence.

Gq-Gs pathways work in the adhesive papillae for metamorphosis

The above results showed that activation of the Gq and Gs pathways are the key events in initiating metamorphosis. Gαq is necessary to induce Ca2+ transients in the adhesive papillae, suggesting that the Gq pathway functions in this region. If both Gq and Gs function in the papillae, they should be expressed in the adhesive papilla. The transcriptome analysis of the larval papilla region showed that the genes encoding Gq and Gs pathway proteins are expressed in this region (Figure S1C, Figure S3 and Table S1). Therefore, Gq and Gs could function in the papilla to initiate metamorphosis.

As shown above, theophylline induced metamorphosis without settlement. However, when the papillae were removed from larvae, the average rate of metamorphosis induction by theophylline was reduced from 100% to 30% (Figure 3A-C). This suggests that cAMP elevation in the adhesive papillae is essential for starting metamorphosis. The overexpression of caGαq, caPLCβ1/2/3, caGαs, and bPAC by the PC2 cis element resulted in the initiation of metamorphosis without adhesion. Like theophylline, amputation of the papillae reduced their effects in starting metamorphosis (Figure 3D), confirming that activation of the Gq and Gs pathways in the adhesive papillae triggers metamorphosis. Among these experiments, caPLCβ1/2/3 overexpression was the most sensitive to papilla amputation, suggesting that PLCβ1/2/3 acts specifically in the papillae during metamorphosis.

Gq and Gs pathways are activated in the adhesive papillae to initiate metamorphosis.

(A) A theophylline-treated animal. Metamorphosis initiated.

(B) A papilla-amputated larva treated with theophylline. Metamorphosis did not initiate.

(C) Effect of papilla amputation on theophylline treatment.

(D) Effects of papilla amputation on caGαq, caPLCβ1/2/3, caGαs, and bPAC overexpression.

(E) Measurement of cAMP concentration, as indicated by the fluorescence of Pink Flamindo. Left, the quantification of Pink Flamindo fluorescent intensity is shown for a larva that initiated metamorphosis at the time indicated by an arrowhead. The time of stimulation of the adhesive papillae with a glass rod is denoted by a bar. Right, the position of the region of interest (ROI). See also Figure S4.

If the Gs pathway is activated in the adhesive papillae, a cAMP increase should be observed in this region upon adhesion. We examined this possibility using a fluorescent cAMP indicator called Pink Flamindo (PF)60. After stimulation, the fluorescence of PF in the papillae temporarily decreased on average by 0.83-fold (n=5) from the initial intensity, followed by a gradual increase by 0.0102 per min, to reach 1.17-fold (Figure 3E and Figure S4A). The cAMP reduction and increase respectively started at 35 seconds and 4 min 40 seconds after stimulation on average. Such a reduction and increase in fluorescence intensity was not observed in the adhesive papillae of the larvae that had failed to initiate metamorphosis following stimulation (n=4; Figure S4B). Therefore, increased cAMP in the papillae serves to indicate that sufficient stimuli of adhesion have been received to induce metamorphosis. This strengthens our hypothesis that cAMP accumulation in the adhesive papillae determines the initiation of metamorphosis. Both the Gαs and Gαq knockdowns abolished the temporal cAMP decrease, and reduced the rate of subsequent accumulation (Figure S4C and D), indicating that these two events are dependent on the Gs and Gq pathways. However, neither knockdown abolished the increase in cAMP completely (as shown by the slope scores), suggesting the presence of a mechanism producing cAMP without their activations.

GABA in the adhesive papillae is responsible for metamorphosis

Our previous study demonstrated that GABA is the chief neurotransmitter that induces metamorphosis39. To gain insight into the relationships between the GABA, Gq, and Gs pathways, we addressed whether GABA functions in the adhesive papillae for initiating metamorphosis, similar to Gq and Gs.

The previous studies and our transcriptome data suggest that GAD is expressed in the adhesive papillae61 (Table S1). Moreover, GABA-immunopositive signals have been detected in the papillae23,61. Therefore, papillae can provide GABA to stimulate themselves. Next, we examined whether the adhesive papillae can receive GABA to initiate metamorphosis. We showed that the metabotropic GABA receptor is responsible for the initiation of metamorphosis39. Using whole-mount in situ hybridization, the previous study did not detect the expression of two GABAB receptor (GABABR) genes in the papillae61; however, our transcriptome data detected the low-level expression of three genes encoding GABABR (Table S1) in the papillae, suggesting that adhesive papillae could receive GABA. GABA can induce metamorphosis without adhesion (Figure S5A)39. However, the amputation of adhesive papillae suppressed this activity (Figure S5B-C), suggesting that GABA reception by the papillae is responsible for starting metamorphosis.

The larval brain is the major domain expressing GABABR61. Therefore, it remains possible that GABA signaling in the brain stimulates the papillae in a retrograde manner to initiate metamorphosis. The connectome analyses of the larval nervous system did not suggest a nerve that inputs into the papillae from the brain62,63; however, retrograde stimulation would be possible through the volume transmission of a signaling molecule. To examine this possibility, we isolated larval fragments anterior to the brain (Figure S5D), followed by the administration of GABA and theophylline. Through the application of these chemicals, anterior fragments exhibited increased clarity, elongation, and retraction of papillae, which are the signatures of metamorphosis (Figure S5E-G). Because their responses to GABA were weaker than those observed in the theophylline treatment, we further examined whether the anterior fragments can respond to GABA by monitoring the expression of GABA-responsive genes. By comparing expression levels between control and GABA-administered larvae, we made a list of the genes upregulated by GABA in the papillae (Table S2). Among them, we compared the expression levels of two genes (KY21.Chr5.240 and KY21.Chr8.489) between GABA-administered and control anterior fragments. GABA-treated anterior fragments expressed these genes at levels at least three times higher than controls (Figure S5H). We concluded that GABA is secreted from and stimulates the papillae to start metamorphosis.

Gs/cAMP is downstream of the Gq pathway

Our next question is: What are the upstream or downstream relationships between GABA, Gq/Ca2+, and Gs/cAMP in the adhesive papillae? We first examined the relationships between the Gq and Gs pathways. Theophylline ameliorated metamorphosis-failed phenotypes in Gαq, PLCβ, and IP3R knockdowns (Figure 4A-E). Moreover, caGαs overexpression in the nervous system significantly, although not perfectly, ameliorated the effect of Gαq MO (Figure 4F). These results suggest that the Gq pathway is upstream of the Gs/cAMP pathway. Gαs knockdown larvae exhibited Ca2+ transients in the adhesive papillae upon stimulation (Figure 4G). Because this Ca2+ transient is Gq-dependent (Figure 1O), this result confirms that the Gs pathway is downstream of the Gq-dependent Ca2+ increase.

Relationships between Gq and Gs pathways in metamorphosis.

(A) A Gαq knockdown larva.

(B) Theophylline ameliorated the effect of Gαq knockdown.

(C) Effect of theophylline on Gαq knockdown.

(D) Effect of theophylline on PLCβ knockdown. In this experiment, PLCβ1/2/3 and PLCβ4 were knocked down simultaneously.

(E) Effect of theophylline on IP3R knockdown.

(F) Effect of caGαs overexpression on Gαq knockdown.

(G) Gαs is not necessary for Ca2+ transient in the adhesive papillae (arrow) upon adhesion. The percentage and number represent the rate of animals showing Ca2+ transient in the papillae.

(H) A Gαq knockdown and Kaede-overexpressed larva.

(I) A Gαq knockdown and caPLCβ1/2/3-overexpressed larva.

(J) A Gαs knockdown and caPLCβ1/2/3-overexpressed larva.

(K) Effect of caPLCβ1/2/3 overexpression on Gαq knockdown.

(L) Effect of caPLCβ1/2/3 overexpression on Gαs knockdown.

(M) Effect of caGαq overexpression on Gαs knockdown.

(N) Effect of theophylline treatment on Gαs knockdown.

To gain further insight into the epistatic order of the Gq and Gs pathways, we overexpressed constitutive active forms of Gq pathway proteins in Gαs morphants with the PC2 cis element. If Gq is upstream of the Gs pathway, forced activation of the Gq pathway by caGαq or caPLCβ1/2/3 would not ameliorate the Gαs MO effect. Indeed, caPLCβ1/2/3, which rescued Gαq morphants well, failed to rescue Gαs morphants (Figure 4H-L). In contrast, caGαq significantly ameliorated the metamorphosis-failed phenocopies of Gαs morphants (Figure 4M), contradicting the hypothesis that Gq is upstream of the Gs pathway. One possibility is that the Gq pathway stimulates cAMP synthesis through a massive Ca2+ increase and protein kinase C (PKC) activation64,65. Indeed, Ciona larvae can synthesize cAMP without Gs, as suggested by the imaging of Pink Flamindo (Figure S4C), and by the partial rescue of Gαs morphants by theophylline treatment (Figure 4N). These data suggest that Gαs-independent AC could be a target of the Gq pathway. We concluded that the Gq pathway is upstream of the Gs pathway in the signaling cascade initiating Ciona metamorphosis.

Interaction between GABA and Gq pathways

We next investigated the relationships between the GABA and Gq/Gs pathways. As our previous study showed, GABA pathway knockdown by GAD or GABABR1 MO disrupted the initiation of metamorphosis (Figure 5A)39. Overexpression of caPLCβ1/2/3 and theophylline treatment ameliorated the metamorphosis-failed phenocopies of GAD/GABABR1 morphants (Figure 5B-D). These results suggest that the GABA pathway is upstream of the Gq and Gs pathways. We observed Ca2+ transients in the adhesive papillae of GAD knockdown larvae. GAD morphants rarely exhibited a Ca2+ increase after stimulation (Figure 5E). Because the Ca2+ transient in the papillae is Gq-dependent, this result confirms that the GABA pathway is upstream of, or at least in parallel with, the Gq pathway.

GABA functions in the adhesive papillae to induce metamorphosis.

(A) A GABABR1 knockdown and mCherry-overexpressed larva.

(B) A GABABR1 knockdown and caPLCβ1/2/3-overexpressed larva. Metamorphosis initiated.

(C) Effect of caPLCβ1/2/3 overexpression on GABABR1 knockdown.

(D) Effect of theophylline on GABABR1 and GAD knockdowns.

(E) GABA is necessary for Ca2+ transient in the adhesive papillae (arrow). The percentage and number represent the rate of animals showing Ca2+ transient in the papillae.

(F) Effect of GABA on Gαq knockdown.

(G) Effect of GABA on PLCβ knockdown.

(H) Effect of GABA on Gαs knockdown. See also Figure S5.

However, puzzling results were obtained when we administered Gαq- and Gαs-knockdown larvae with GABA. If GABA is upstream of the Gq and Gs pathways, this chemical does not induce metamorphosis when the Gq or Gs pathway is disrupted. However, GABA partially but significantly ameliorated the metamorphosis-failed phenocopies of Gαq, PLCβ, and Gαs morphants (Figure 5F-H). Among the three morphants, GABA achieved more effective rescue in Gas knockdowns than Gαq or PLCβ. These results could be explained by assuming enhancement of the Gq pathway by GABA through PLCβ and another GABA-mediated metamorphic pathway bypassing Gq components. These possibilities are examined in the next section.

Contribution of Gi to metamorphosis

The function of GABA as the potential upstream factor of Gq could be easily explained if GABABR is coupled with Gq. Although a few studies suggest this possibility66, Gi is regarded as the major G-protein coupled with GABABR67. A previous study demonstrated that the gene encoding typical Ciona Gαi is strongly expressed throughout the entire larval nervous system, including in adhesive papillae68, and our RNA-seq confirmed this (Figure S1C and Table S1). The knockdown of this gene (Gαi) exhibited a moderate (although statistically significant) reduction of metamorphosis rate (Figure S6A), suggesting the presence of another Gαi regulating metamorphosis.

Indeed, the dvGαi_Chr2 knockdown larvae failed to initiate metamorphosis (Figure 6A-C), and this phenocopy was rescued by theophylline and GABA (Figure 6D), suggesting that dvGαi_Chr2 is chiefly necessary for metamorphosis. The knockdown of dvGαi_Chr4 resulted in a slight reduction of the metamorphosis rate (Figure S6B), suggesting the supportive role of this Gαi in the metamorphosis initiation. These data strengthen the hypothesis that GABABR regulates metamorphosis through Gi. The rescue of dvGαi_Chr2 morphants by GABA might be attributable to a redundant function of the typical Gαi and dvGαi_Chr4 proteins in the papillae; however, we could not fully address this hypothesis due to a technical limitation. That is, the simultaneous knockdown of dvGαi_Chr2 and dvGαi_Chr4 did not suppress the amelioration of metamorphosis by GABA (Figure S6C). This might have been due to the presence of typical Gαi. The simultaneous knockdown of dvGαi_Chr2 and typical Gai resulted in malformation of the body (Figure S6D), and we could not examine their responsiveness to GABA.

Gi is required for metamorphosis initiation.

(A) A control animal injected with STD MO.

(B) A dvGαi_Chr2 knockdown larva.

(C) A dvGαi_Chr2 knockdown and theophylline-treated animal.

(D) Effects of GABA and theophylline on dvGαi_Chr2 knockdown.

(E) Measurement of cAMP concentration in the dvGαi_Chr2 knockdown larvae. The red and pink graphs represent the averaged Pink Flamindo fluorescent intensity and the standard deviations, respectively. The number in the parentheses indicates the number of examined larvae. The red graph is approximated to the linear graph to calculate its slope, which is shown as Δ. See also Figure S4.

(F) Effects of wtGαi and wtGαs overexpression.

(G) Effect of GABABR1 and Gαq simultaneous knockdown.

(H) Effects of caPLCβ1/2/3 and caGαs overexpressions on GABABR1 and Gαq simultaneous knockdown.

(I) GIRK knockdown promoted metamorphosis initiation. See also Figure S6.

We further explored the involvement of Gi upon metamorphosis as the upstream factor of the Gq-Gs pathways. Knockdown of dvGαi_Chr2 resulted in the abolishment of temporal cAMP decrease and its subsequent accumulation upon adhesion (Figure 6E). Because these cAMP fluctuations are Gq- and Gs-dependent (Figure S4C-D), their dependence on dvGai_Chr2 supports that Gi is upstream of the Gq-Gs pathways. Gi is known to activate PLCβ through the Gβγi complex69,70.

Released βγ is inactivated by overexpressing normal (usually GDP-bound) Gα subunits because GDP-Gα quenches the βγ complex71. Overexpressing wild-type Gαi and Gαs in the nervous system significantly suppressed metamorphosis (Figure 6F), suggesting that the activated βγ complex is necessary for initiating metamorphosis. To confirm that the wild-type Gαi exerts its effect through sequestration of the βγ complex, we overexpressed a dominant-negative form of Gαi (dnGαi)72,73 which has reduced GDP/GTP affinity while maintaining βγ binding activity. dnGαi significantly reduced the occurrence of metamorphosis (Figure S6E), suggesting that the negative effect of Gαi on metamorphosis occurs through interaction with the βγ complex.

If Gβγ-mediated PLCβ activation is used in Ciona metamorphosis, PLCβ receives two independent inputs (Gαq and Gβγi) for its activation. These pathways could compensate for each other. We noticed that the MOs for GABA pathway genes and the Gαq MO did not disrupt metamorphosis as strongly as the Gαs MO (Figure 1E-F; 5C-D). The compensatory role could explain this phenomenon. Indeed, the simultaneous knockdown of GABABR1 and Gαq resulted in a strong impairment of metamorphosis (Figure 6G), and these morphants initiated metamorphosis by overexpression of caPLCβ1/2/3 or caGαs (Figure 6H).

If the role of the GABA/Gi pathway relies specifically on PLCβ activation in the mechanism of metamorphosis, GABA administration would not rescue PLCβ morphants. However, GABA weakly but significantly induced metamorphosis in PLCβ1/2/3 plus PLCβ4 double-knockdown larvae (Figure 5G). Therefore, GABA/Gi is likely to activate another pathway that bypasses PLCβ. Gβγi is known to activate group III AC64. Its Ciona counterpart (AC5/6) is expressed in the adhesive papillae (Figure S3B), which could explain the results of the rescue experiments. In addition, there are two other major targets of Gβγi. One is the G-protein-activated inwardly rectifying potassium (GIRK) channel74. Our RNA-seq data on the papillae indicated the expression of two GIRK channel genes (Table S2). The GIRK channel negatively regulates the excitation of neurons through hyperpolarization. If the metamorphosis of Ciona is induced by the excitation of papilla neurons as suggested previously18,26, the GIRK channel is likely to regulate metamorphosis negatively. Indeed, the knockdown of one GIRK channel gene weakly enhanced the initiation of metamorphosis without settlement (Figure 6I). Although the negative role of the GIRK channel supports the involvement of Gβγi in the pathway of metamorphosis, this does not explain the PLCβ-independent activation of the metamorphic pathway by GABA/Gi.

The third function of Gβγi is to activate MAPK signaling through positive regulation of MEK75. PLCβ does not mediate this pathway. It has been suggested that MAPK signaling is required to induce metamorphosis76,77. To show that MEK activation is necessary for inducing metamorphosis mediated by GABA, we treated adhesion-prevented larvae with GABA plus U0126, a potent MEK1/2 inhibitor repeatedly used in ascidians78,79. U0126 significantly reduced the rate of metamorphosis induction by GABA (Figure S6F).

cAMP is also known to activate the pathway that involves MEK1/280. We found that U0126 antagonized the effect of theophylline on metamorphosis (Figure S6G). Therefore, GABA and cAMP have the same target (MEK1/2) to initiate metamorphosis, suggesting their compensatory function and/or that cAMP synthesis is upregulated by GABA-Gβγi. These bypassing pathways could explain the amelioration of the phenocopies of PLCβ and Gαs morphants by GABA (Figure 5G and H).

The constitutive function of wild-type Gαq

We found that Ciona wild-type Gαq (wtGαq) has constitutive activity. In contrast to wtGαs and wtGαi, overexpression of wtGαq did not arrest metamorphosis (Figure S7A). Rather, its overexpression induced metamorphosis without settlement (Figure S7B). The constitutive wtGαq activity was confirmed by the significant rescue of Gαs morphants (Figure S7C). As mentioned above, caGαq also has these activities (Figure 1L). caGαq-overexpressed larvae had a somewhat rounder trunk shape, abnormal tail bending, and frequent tail twitching, perhaps due to a constitutive increase in the cytosolic Ca2+ concentration (Figure S7D). Overexpression of wtGαq did not show such abnormalities (Figure S7E), suggesting that wtGαq has milder activity than, or works through a different mechanism from, caGαq. To gain further insight into the constitutive effect of wtGαq, we constructed a dominant-negative form of Gαq72 that has reduced GDP/GTP binding activity while maintaining βγ binding activity. Overexpression of dnGαq weakly inhibited metamorphosis (Figure S7F), suggesting that GDP/GTP exchange is important for the constitutive activity of wtGαq.

Discussion

Among chordates, the tunicate ascidian is the only group that exhibits a sessile lifestyle at the adult stage. Understanding how ascidians acquired the mechanism to metamorphose into sessile adults is important for elucidating the evolution of chordates81,82. Through extensive molecular, physiological, and pharmacological analyses, we identified signaling molecules responsible for the initiation of metamorphosis of the ascidian Ciona. Combining the results with previous knowledge, we described a schematic that best represents the cascades running in the adhesive papillae upon adhesion to initiate metamorphosis (Figure 7). This working hypothesis will be the basis for future research on ascidian/tunicate metamorphosis. The hypothesis explains the characteristics of Ciona metamorphosis, the mechanisms of which have not yet been elucidated. The genes, proteins, and signaling pathways in this schematic are essential targets for elucidating how the ancestor of ascidians acquired the metamorphosis system peculiar to this group during evolution.

Schematic illustration of the signaling cascades for initiating Ciona metamorphosis.

AC, adenylate cyclase; ER, endoplasmic reticulum; GPCR, G-protein coupled receptor; PKA, protein kinase A.

Mechanisms measuring duration and strength of adhesions

One mystery of Ciona metamorphosis is that its initiation requires continuous adhesion with the adhesive organ for a certain period. This requirement is suggested for the faithful achievement of metamorphosis only when adhesion is firm enough to allow the transition out of the free-living larval stage. Our previous study suggested that approximately 30 min of adhesion is necessary31. A shorter duration of adhesion does not trigger metamorphosis. Larvae that experience short-term adhesion (such as detaching before 30 min) need to adhere again for 30 min to initiate metamorphosis, suggesting that the experience of temporal adhesion is erased. Our recent study showed that the force given to the papillae by adhesion also affects metamorphosis initiation: weaker force extends the time requirement26. Therefore, Ciona larvae likely possess a mechanism that somehow measures the duration and strength of adhesion and the ability to cancel transient excitations of the adhesive organ. Because fewer than 300 neurons constitute the larval nervous system of Ciona62,63,83, how the larva measures the strength and duration of adhesion with its simple nervous system is an important question to address the mechanisms of metamorphosis.

This study showed that cAMP is an essential second messenger molecule for triggering metamorphosis. Because PDE constitutively degrades cAMP, accumulation of this molecule requires persistent and/or strong activation of the metamorphic pathway that overcomes PDE activity. In other words, the duration and strength of adhesion could be converted into the quantity of cAMP, and when its quantity reaches a threshold, metamorphosis can be initiated. The time between the start of adhesion and metamorphosis initiation may be the period necessary to accumulate sufficient cAMP. This “cAMP timer” mechanism does not demand a complicated neuronal network system and could work in the simple nervous system of Ciona. Prior to triggering the cAMP timer, Ciona temporarily decreases cAMP quantity (Figure 3E and S4A). This reduction is also associated with metamorphosis initiation, suggesting that Ciona does not simply sense the cAMP increase. There are two plausible functions of this reduction. One is guaranteeing that the requirement of long-term adhesion will be met until a sufficient quantity of cAMP is accumulated, irrespective of the initial quantity of cAMP. The other one is multiplying the degree of cAMP increase. By reducing the initial value, larvae can increase the amount of this molecule several fold, even though the maximum quantity remains the same. Such an elevated difference in cAMP quantity may be adopted to ensure that the timing of metamorphosis initiation is appropriate using this ubiquitous molecule.

G-protein signaling relay for metamorphosis

This study showed that three major G-proteins are involved in the initiation of metamorphosis (Figure 7). This complicated mechanism stands in contrast to the simple determination of metamorphosis initiation by cAMP quantity. What factor promoted the ancestor of ascidians to acquire this complicated signaling network of metamorphosis? We suspect that incorporating multiple components increases the chance of generating the crosstalk between molecules, providing rigidity and flexibility to the system. Particularly, the involvement of Gi may have been important for achieving metamorphosis at the correct time and condition. GABA and Gi have inhibitory and excitatory functions84,85. Ciona larvae must ignore transient adhesion or stimulation on the papilla, which may occur frequently during swimming by colliding with an object or by strong water flows. At the same time, larvae must maintain sensing ability and start metamorphosis when they encounter an appropriate stimulus. Through the GIRK channel and Gαi, GABA can suppress the excitation and cAMP accumulation in the papilla. The GIRK channel could increase the excitation threshold of papilla neurons, allowing them to ignore weak stimuli. The transient reduction of cAMP at the onset of adhesion could be partly explained by the inhibitory function of Gαi on ACs. Not only Gi, but Gq and Gs are also necessary for this cAMP reduction. PDE1 is known to be activated by Ca2+/calmodulin86. Because PDE1 is abundantly expressed in the papillae (Figure S3B), the increase in Ca2+ upon adhesion by Gi and Gq could also trigger sudden cAMP reduction through PDE1. The contribution of Gs to cAMP reduction is surprising because the general role of this complex is to increase cAMP. However, it is known that an important target of the Gs pathway for the termination of cAMP synthesis is PDE86. The involvement of three G-protein pathways suggests that the acute cAMP reduction is a special event that is distinct from its constitutive degradation.

Together with its inhibitory function, GABA can activate the Gq pathway, which promotes metamorphosis through Gβγi. Group III AC and MEK could also be the targets of Gβγi. When adhesion is long enough, their activations somehow overcome the inhibitory functions of GABA. The GABA pathway is an ideal player that fulfills multiple requirements in the regulation of metamorphosis through its excitatory and inhibitory functions. We suspect that papillae themselves are the source of GABA for metamorphosis initiation. Because VIAAT/VGAT is not expressed in the papillae, the secretion of GABA from the papilla neurons requires an atypical mechanism. GABA is detected in the cell body of the neurons23. The release of cytosolic GABA may be triggered by the reception of mechanical stimuli using the TRP channel26. A VIAAT-independent cytosolic GABA release is observed in pancreatic beta cells87. Curiously, papilla neurons express transcription factor Islet88, which is homologous to the vertebrate transcription factor responsible for β cell function. The atypical secretion of GABA might depend on a transcription factor like Islet shared between Ciona papilla neurons and vertebrate beta cells.

A recent study19 provided several lines of evidence suggesting that PNs can serve as the sensors of several chemicals in addition to mechanical stimuli. This finding and our model could be mutually related because these chemicals could modify Ca2+ and cAMP production. G protein signaling allows Ciona to reflect various environmental stimuli to initiate metamorphosis either mechanically or chemically according to the situation.

GPCRs for initiating metamorphosis and atypical Gαq activity

Future studies need to specify the mechanisms underlying Gαi, Gαq, and Gαs activation as well as the interaction between them more precisely. The activation of trimeric G-proteins relies on GPCRs, suggesting the presence of the receptors coupled with Gi, Gq, and Gs for metamorphosis initiation (Figure 7). Among them, Gi is likely to be coupled with GABABR; however, the expression of GABABR in the papillae is weak, and their interaction needs to be clarified in the future. We found that wild-type Gαq exhibits constitutive activity that does not require activation through adhesion. The constitutive Gαq activity requires GTP binding. If a GPCR owns this Gαq activation, its ligand should be supplied constitutively. Wild-type Ciona does not initiate metamorphosis without adhesion even though Gαq is expressed in the adhesive papillae, suggesting that the quantity of Gαq is strictly regulated to prevent autonomous metamorphosis in normal conditions.

We did not identify the GPCR coupled with Gs in the metamorphic mechanism. We suspect that GnRH receptors (GnRHRs) are strong candidates for this role. Our previous studies showed that GnRHs can induce metamorphosis39,40 as a downstream factor of GABA. Among the four genes encoding GnRHR proteins, GnRHR1 and GnRHR2 are expressed in the adhesive papillae89. These GnRHRs stimulate cAMP signaling, and the ligand GnRHs for GnRHR1 are also expressed in the papillae89–91. Therefore, GnRHs could be secreted from adhesive papillae as downstream molecules of GABA and stimulate the Gs pathway through GnRHR1. Future studies targeting the regulation of the secretion and reception of GnRH peptides after stimulation of papillae will improve our understanding of the signaling cascade conducting Ciona metamorphosis.

Understanding whether the G-protein signaling relay occurs in cell-autonomous or non-autonomous fashions is crucial for deepening our understanding of Ciona metamorphosis. Each papilla comprises approximately 20 cells, including four papilla neurons (PNs). We showed that activation of Gq and Gs pathways in PNs is sufficient for initiating metamorphosis. This coincides with the previous reports26,27 that the loss of PNs by disrupting the POUIV transcription factor caused the complete arrest of metamorphosis. However, we do not think PNs are the only cells that activate G-proteins, because Ca2+ and cAMP imaging showed upregulation of fluorescence in the entire papillae18,26. A recent study also reports Ca2+ transient in another cell type than PNs19. Other papilla cells are likely responsible for activating G-proteins by directly responding to adhesion and by receiving signal input from adjacent cells, thereby enhancing the signal strength to accumulate a sufficient quantity of cAMP in the PNs to initiate metamorphosis. In this study, technical limitations prevented us from characterizing cells exhibiting increases in Ca2+ and cAMP; we need to address this question in future studies.

Evolutionary implications

Sessile or benthic marine invertebrates lose locomotive activity when they undergo metamorphosis triggered by adhesion to a substratum92. These animals are suspected of having a system to control the adhesion state, allowing them to repeat attachment and detachment before meeting the appropriate conditions for metamorphosis. For example, barnacle larvae “walk” on the substratum before adhering firmly by secreting cement93. Like Ciona, the larvae of some sessile/benthic animals may have a system to initiate metamorphosis only when appropriate adhesion is provided, while erasing stimuli from transient and inappropriate adhesion. GABA serves as the metamorphosis inducer of some benthic invertebrates, including mollusks and echinoderms94–96. Moreover, GPCRs are implicated as the mediators of settlement and metamorphosis induction in hydrozoans, mollusks, and barnacles94,97,98. In the abalone and barnacle, the dual use of Gq and Gs pathways in metamorphosis has been suggested99,100. Supported by these shared features in the metamorphic mechanisms, our working hypothesis about the initiation of Ciona metamorphosis (Figure 7) will serve as a cue to elucidate how marine benthic invertebrates regulate their metamorphosis.

Materials and methods

Animals

Ciona intestinalis

Type A wild types collected from Onagawa Bay (Miyagi, Japan) and Onahama Bay (Fukushima, Japan) were cultivated in closed colonies by the staff of the National BioResource Project, Japan. They were kept under a constant light condition to prevent gamete release. Eggs and sperm were collected surgically from gonadal ducts, and insemination was carried out in dishes. To prevent larvae from initiating metamorphosis, the tail’s posterior half was manually cut with a scalpel. Removing the tail prevents larvae from swimming efficiently, and these larvae are usually unable to adhere to a substrate. These adhesion-prevented larvae do not metamorphose, since adhesion is required to initiate metamorphosis. Larvae developed from dechorionated eggs were cultured on a 2% agar-coated dish after tail amputation to prevent metamorphosis. Because larvae stick to plastics, tail amputation is insufficient to avoid their adhesion to the culture dish.

Pharmacological treatment

Larvae or larval anterior fragments isolated with a scalpel were administered overnight with 700 μM GABA (Fujifilm Wako #010-02441) or 1 mM theophylline (Sigma-Aldrich #T1633) dissolved in seawater, or 4 μM U0126 (Promega #V1121) dissolved in DMSO.

Plasmids

The open reading frame (ORF) of Kaede was removed from pSPCiPC2Kaede by inverse PCR with PrimeStar GxL DNA polymerase (Takara-bio #R050). The ORFs of Gαq, Gαs, Gαi, and PLCβ1/2/3 were PCR amplified using full-insert cDNAs101. These PCR fragments were fused with the In-Fusion HD Cloning kit (Clontech #639650). The region encoding XY linker51 was deleted from PLCβ1/2/3 cDNA by inverse PCR to create a constitutively active form. The mutations were introduced by inverse PCR to create constitutive active or dominant-negative forms of Gα cDNAs. The introduced mutations are as follows: caGαq, corresponding to Q223L, abolishes GTPase activity50; caGαs, corresponding to Q227L, abolishes GTPase activity55; dnGαi, corresponding to S47C72,73; and dnGαq, corresponding to S54N72. The ORF of bPAC59 was PCR amplified and inserted between the BamHI and EcoRI restriction sites of pSP-Kaede102 to create pSPbPAC. The PCR-amplified PC2 cis-element54 was inserted into the BamHI site of pSPbPAC. The plasmid DNAs were linearized by a restriction enzyme and purified with the Qiaquick Gel Extraction Kit (Qiagen #28706) before microinjection. The ORF of Pink Flamindo60 was PCR amplified and inserted into the EcoRV restriction sites of pBS-HTB103,104 to create pHTBPinkFlamindo. The plasmids pHTBGCaMP826 and pHTBPinkFlamindo were linearized using XhoI for subsequent in vitro synthesis of GCaMP8 and Pink Flamindo mRNA, respectively. GCaMP8 mRNA was synthesized with the MEGAscript T3 kit (Thermo Fisher Scientific #AM1338), the Poly (A) Tailing kit (Thermo Fisher Scientific #AM1350), and the Cap structure analog (New England Biolabs #S1404). Pink Flamindo mRNA was synthesized with the mMESSAGE mMACHINE T3 Transcription kit (Thermo Fisher Scientific #AM1348).

Microinjection

Unfertilized eggs were dechorionated in sterilized seawater containing 1% sodium thioglycolate (Fujifilm Wako #590-11762) and 0.05% actinase E (Kaken Pharmaceutical #650164) as previously described105. The microinjection solution included 2 mg/ml of Fast Green FCF (Fujifilm Wako #061-00031), 0.5-1.0 mM of morpholino oligonucleotide(s) (MOs)106, 5 ng/ul of plasmids, and/or 1 mg/ml of GCaMP8 mRNA for imaging26. For cAMP imaging, 1 mg/ml Pink Flamindo mRNA was dissolved in water without Fast Green, since this chemical emits fluorescence. Microinjected unfertilized eggs were inseminated in a gelatin-coated plastic dish. After the seawater was exchanged to remove excess sperm, the fertilized eggs were cultured at 18 °C or 20 °C overnight until the larval stage. MOs are listed below. Standard control MO, 5’- CCTCTTACCTCAGTTACAATTTATA-3’; GAD ATG MO39, 5’-ACCTCCAAGCCGATTGTTTCTGCAT-3’; GABABR1 ATG MO39, 5’- GCTTACGACTTTACATAACCTTACA-3’; Gαq ATG MO, 5’- GGCATATTTGTGACTATAATGACG-3’; Gαs ATG MO, 5’- AAAGCAACCCATTGGCATTATCGAC-3’; PLCβ1/2/3 splicing MO, 5’- GTGTTACTTACGCTTTCTCTA-3’; PLCβ4 splicing MO, 5’- AACCACCAACCACCAACCTTTTG-3’; IP3R splicing MO, 5’- AATGATGGTTTAAAATTGCCACCTG-3’; Gαi ATG MO, 5’- GTGGAGACTGTGCAACCCATGATTC-3’; dvGai_Chr2 ATG MO, 5’- CCATCTTGAGTAATCCAGGCTTTTA-3’; dvGai_Chr4 ATG MO, 5’- CATGGTCAGCGGTTTACAAAGTATT-3’; GIRK channel ATG MO, 5’- TCTGCTGGTTCAGTAATAGACATAG-3’.

Photographing and imaging

Photographs were taken with an AxioImager Z1 and AxioObserver Z1 (Carl Zeiss). The images were treated with AxioVision Rel.4.6 or Zen (Carl Zeiss) and Photoshop 2021 (Adobe) software. Imaging of the Ca2+ transient was carried out according to previous reports18,26. The tails of GCaMP8 mRNA-injected larvae were removed by a scalpel at 15-17 hours after fertilization (hpf). At 30-40 hpf, the larvae were mounted onto a plastic dish under an M165FC stereo microscope (Leica Microsystems), and their adhesive papillae were stimulated by a glass rod using a three-axis coarse manipulator M-152 (Narishige Japan). Immediately after the tip of the rod had contact with the papillae, visible light was shut off and the fluorescence of GCaMP8 was taken with a DFC 310FX digital camera (Leica), Las version 4.12 (Leica), and the movie-capturing function of Windows 10. Images were treated with Photoshop 2021.

Imaging of cAMP was taken by confocal laser scanning microscopy (FV1000, Olympus). The movies were treated with ImageJ (National Institutes of Health). The Pink Flamindo mRNA-injected larvae were immobilized on Poly L lysine-coated glass-bottom dishes at 20-21 hpf, and their adhesive papillae were stimulated around 25 hpf. For the stimulation, an electric manipulator MM-89 (Narishige) and hydraulic micromanipulator MMO-202ND (Narishige) were used.

RNA sequencing

Larval anterior portions, including adhesive papillae and the other trunk region, were separately isolated by a scalpel around 22-25 hpf at 18°C and collected in Isogen (Nippon Gene #317-02503) soon after isolation. To identify GABA-responsive genes, half of the tail was amputated from larvae around 21-24 hpf, followed by GABA administration for 2 hours before treatment with Isogen. In each experiment, approximately 50 fragments were collected, and two biological replicates were taken. Total RNA was isolated according to the manufacturer’s instructions. Ribosomal RNA was depleted from the total RNA using the NEBNext rRNA Depletion kit (New England Biolabs #E6310), followed by the conversion of the remaining RNA into an Illumina sequencing library using the NEBNext Ultra Directional RNA Library Prep kit (New England Biolabs #E7760). Following library preparation, validation was performed using the Bioanalyzer system (Agilent Technologies) to assess size distribution and concentration. Subsequently, sequencing was carried out on the NextSeq 500 platform (Illumina) employing the paired-end 36-base read option. Reads were mapped on the Ciona intestinalis Type A reference genome (HT model; http://ghost.zool.kyoto-u.ac.jp/default_ht.html) and quantified using CLC Genomic Workbench version 22.0 (Qiagen). RNA-seq data sets were deposited in the NCBI Gene Expression Omnibus (GEO) under accession numbers SAMN40712866 to SAMN40712873, which will appear online upon publication of this paper.

The read counts were normalized by calculating the number of reads per kilobase per million (RPKM) for each transcript in each sample using CLC Genomic Workbench version 22.0 (Qiagen). Eight candidate genes showing a minimum 1.5-fold increase following GABA administration, along with notably higher expression in the papillae in comparison to the trunk (an approximately 4-fold increase in a sample), were chosen for quantitative PCR analysis to assess their response to GABA.

Quantitative PCR

Total RNA was isolated from the anterior tips of larvae treated with the chemical using Isogen (Nippon Gene), following the manufacturer’s instructions. After isopropanol extraction, genomic DNA removal and reverse transcription were carried out using the PrimeScript™ RT reagent kit with gDNA Eraser (Takara Bio #RR047). Quantitative RT-PCR (qRT-PCR) was done using a SYBR Premix Ex TaqII Dimer Eraser (Takara Bio #RR091) and a Thermal Cycler Dice Real Time System III (Takara Bio). The gene encoding elongation factor 1α (EF1α) was used to normalize RNA quantity according to a previous study104. qPCR primers are listed below. KY21.Chr5.240, 5’- TCTTCTCAAAGTTGCACATTCC-3’ and 5’-CAGCAGCAACCAAACGATAAAC-3’; KY21.Chr8.489, 5’- CAATGCAACTTTGACTGCATAC-3’ and 5’- TCCAAACTGCATTCCACATATC-3’; EF1α, 5’-CATGTCACGGACAGCGAAACG- 3’ and 5’- CAATGTGTGTTGAGGCATTCCAAG-3’.

Statistical analysis

Differences between conditions were evaluated by Fisher’s exact test. Statistical analyses and most graph visualizations were conducted utilizing R x64.4.1.2 and RStudio software.

Phylogenetic analysis

We performed a tBlastn search107 against the HT gene model of Ciona44 using the amino acid sequences of human PDE, PLCβ, and Ciona Gα proteins as queries, followed by a reciprocal Blast search. We aligned the amino acid sequences from the HDc domain of PDE, the EF-hand_like, PLCXc, PLCYc, and C2 domains of PLCβ, and the whole open reading frame of Gα proteins using the M-COFFEE program108,109. After the removal of unnecessary residues, maximum likelihood trees were constructed using MEGA software version 11110, employing the WAG amino acid substitution matrix111. The trees were assessed with 1,000 bootstrap pseudoreplicates.